---
title: 'How profvis helped give a 2x speedup by changing a single function call (answer:
  data frames can be really slow)'
author: Collin Erickson
date: '2017-11-24'
slug: how-profvis-helped-give-a-2x-speedup-by-changing-a-single-function-call-answer-data-frames-can-be-really-slow
categories: [r]
tags: [r]
---

```{r setup0, echo=FALSE}
set.seed(0)
```


Recently I've been running computationally intensive simulations that take
up to six hours each.
This is prohibitively slow, and is assuredly suboptimal since I coded it all,
so I decided to do a time profile evaluation
to find the bottlenecks in the code.
This is easily done in R using the eponymous function from the package `profvis`.

The following example shows how to use `profvis`.
A simple function is created, which has two parts:
the first is a function that runs a loop 
(functions show up better in `profvis` than just code),
the second part is sleeping for 0.4 seconds.
Then it is evaluated inside of `profvis()`.
Displaying the result shows how long each section of the code
took to run.
The profile shows that a little over half the time is spent
in the looping function, while the rest is sleeping.

```{r}
slow_function <- function(n) {
  tf1 <- function() {
    for (x in 1:n) {
      sinx <- sin(x)
      cosx <- cos(x)
      sqrtx <- sqrt(x)
      expx <- exp(x)
    }
  }
  tf1()
  Sys.sleep(.4)
  n
}

slow_function(10)
```

```{r, warning=FALSE}
library(profvis)
pv <- profvis(slow_function(1e6))
pv
```

## My problem

The problem I was working on 
involved modeling a function using a Gaussian process (GP) model,
then using the GP to iteratively select new points to evaluate and 
add to the model.

Each step requires using the GP to get a predicted mean and standard
error at many different points.
I was passing a matrix whose rows are the points to evaluate these predictions at.
The matrix had about 10000 rows, so it was quite large.
To speed up some of the matrix calculations,
I split the function evaluation so it evaluated 50 rows at a time,
which was about the optimal number of rows to use 
(the code wasn't optimized or I wouldn't have had to do this).
After evaluating all groups,
the results from each group were combined back into
a single large data frame.

I ran a time profile on my function and found that
a lot of time was spent creating data frames for each
of the groups before returning.
I didn't need them in data frames then,
I could wait until the final return.
Using a speed test, I determined that lists are much faster.
By making this small replacement, changing `data.frame` to `list`
I cut the run time in half.

## Demonstration

Below is a simplified version of my problem.
In `f1` each element of `x` is evaluated by `f2`.
In `f2` a data.frame is returned with two values,
that are saved back in `f1`.
This code should look useless and poorly optimized,
but it demonstrates what I was working with.


```{r}
f1 <- function(x) {
  out1 <- numeric(length(x))
  out2 <- numeric(length(x))
  for (i in 1:x) {
    y <- f2(i)
    out1[i] <- y$one
    out2[i] <- y$two
  }
  data.frame(one=out1, two=out2)
}
f2 <- function(i) {
  data.frame(one=i, two=i^2)
}
```

Now we can run `profvis` on this to see where the bottleneck is.

```{r}
pf2 <- profvis(f1(1e4))
pf2
```

In the third row up from the bottom, we see that
`data.frame` in `f2` is taking a lot of time.
It turns out that this is specific to using data.frames,
and that lists or returning a numeric of length two is much faster.
In my case I was returning vectors and using a list or `cbind` of the
vectors was better than a data frame.
To see how the speeds of list and data.frame compare, we can use
`microbenchmark` for combining three vectors.

```{r}
c1 <- 1:100
c2 <- runif(100)
c3 <- sample(1:100)
microbenchmark::microbenchmark(list(c1=c1,c2=c2,c3=c3), 
                               data.frame(c1=c1,c2=c2,c3=c3),
                               cbind(c1=c1,c2=c2,c3=c3))
```

From here we can see that lists are almost instantaneous,
and data frames are really slow compared to lists,
with cbind much closer to lists in speed.


## Conclusion

This post won't be very useful to many people,
but the key two points are:

* `profvis` is an important package and function
for code optimization.

* Data frames are much slower to work with than lists.
