---
title: Trying the R package ggrough
author: Collin Erickson
date: '2018-10-20'
slug: trying-the-r-package-ggrough
categories: []
tags: []
---

I recently came across the R package `ggrough` after seeing
[this tweet](https://twitter.com/dataandme/status/1050931390068547584)
while checking the #rstats topic on Twitter.
It makes plots that look like they were haphazardly sketched.
I figured I'd give it a try to see if I could get it to work.

You can find it on GitHub [here](https://github.com/xvrdm/ggrough).
The package is still in progress and has many problems.

## Set up

First you need to install `ggrough`.

```{r, eval=FALSE}
#install.packages("devtools")
devtools::install_github("xvrdm/ggrough")
```

I also had to install showtext, which I also got from GitHub.

```{r, eval=FALSE}
devtools::install_github("yixuan/showtext")
```

## Getting data

I'm going to try data other than was included in the
example on the GitHub page (`mtcars`).
I found the
[Ballarat pet dataset](https://data.gov.au/dataset/ballarat-registered-dogs-cats/resource/96c5a6de-2f89-4cf5-9b90-159e5be97c31)
from the
[Google Dataset Search](https://toolbox.google.com/datasetsearch).


```{r}
pets <- read.csv("https://data.gov.au/dataset/c916481b-6a03-45fb-a651-7d294e8fd6b8/resource/96c5a6de-2f89-4cf5-9b90-159e5be97c31/download/ballaratregisteredanimals.csv")
```

```{r}
str(pets)
```

## Using `ggrough`

Load the libraries

```{r}
library(ggplot2)
library(ggrough)
library(showtext)
```

First we make the plot with ggplot2.
I'm just going to plot a histogram of the age by the type of animal.

```{r}
# p <- ggplot(data=pets, mapping=aes(x=type)) + geom_bar()
p <- ggplot(data=pets[!is.na(pets$age),], mapping=aes(x=age, fill=type)) + geom_histogram(bins=20) + facet_grid(type ~ .)
p
```


So there's a simple bar chart.
Now we want to make it look rough.

```{r}
options <- list(
  Background=list(roughness=4),
  GeomCol=list(fill_style="solid", bowing=0, roughness=6))
get_rough_chart(p, options, 
                #family = "Homemade Apple", 
                font_size_booster = 1.45)
```

I can't get the image to show up in RStudio.
However, in the RStudio Viewer if I "Export -> Save as Web Page...",
then I can open the html file.
And it shows up after knitting with and viewing in my browser.

<!-- <img src="C://Users//cbe117//Documents//GitHub//website-hugo/petsrough1.html" /> -->

<!-- img above, div below -->

<!-- <div w3-include-html="C://Users//cbe117//Documents//GitHub//website-hugo/petsrough1.html" ></div> -->

<!-- object below -->

<!-- <object type="text/html" data="C://Users//cbe117//Documents//GitHub//website-hugo/petsrough1.html"> -->
<!-- <p>backup content</p> -->
<!-- </object> -->



<!-- iframe below -->

<!-- <iframe src="C://Users//cbe117//Documents//GitHub//website-hugo/petsrough1.html" seamless></iframe> -->


## Another plot

There's a warning on the `ggrough` web page that only one plot
will show up on each page.
I'm going to try a second plot here, giving it a title of "Second plot"
so I can see if it shows up.

```{r}
get_rough_chart(p + ggtitle("Second plot"), options, 
                #family = "Homemade Apple", 
                font_size_booster = 1.45)
```

As warned, it just overwrote the first plot above
and left a white space here.

Also I couldn't the font setting didn't do anything,
and in my other tests I couldn't ever get any font besides the default.

## Conclusion

`ggrough` can make silly looking plots, but still has its issues.
I think it would look better for bar charts where there are spaces between bars.
I would have done more tests and examples, but I can't get the plots
to show up in RStudio and only one will showup in a RMarkdown document.

