---
title: Using active bindings in R6 classes
author: Collin Erickson
date: '2018-01-10'
slug: using-active-bindings-in-r6-classes
categories: []
tags: [R, R6]
---

I've used 
[R6](https://cran.r-project.org/web/packages/R6/vignettes/Introduction.html)
classes in a lot of my R projects in recent years.
They are easier to understand than S3 and S4 objects if
you've ever done programming with objects in other languages.
I also am still not sure how I can have functions act on
data in an S3 object.
I think it's not possible, but I'm really not sure.
And R6 classes are definitely better than RC objects.

Active bindings allow R6 fields to act like functions
(or to have functions look like fields).
My main reason for writing this is to see how slow accessing
active binding fields of R6 classes is.


## What is an R6 class?

Shortly, R6 classes allow you to create objects to store data
and functions that act on the data.
This allows you to create multiple instances of the same
class that will have the same functions available.

For example, below I've defined a `Circle` class,
which I'll explain below.

```{r}
Circle <- R6::R6Class(
  classname = "Circle",
  public = list(
    radius = NULL,
    area = function() {
      pi * self$radius ^ 2
    },
    initialize = function(radius) {
      self$radius <- radius
    }
  )
)
```


You give the variable a name as usual,
but you also give a class name as an argument.
These don't have to be the same.
The first is what you'll type to create a new one,
the latter is what will show up when you call `class`.
Then the second argument is `public`, a list
of items or functions associated with the object.
I have defined a variable `radius` and a function `area`.
I set the radius to `NULL` by default.
The function `area` calculates the area using the `radius`
associated with the object.
Using `self$` is how you can access these.
Finally I also defined an `initialize` function,
which is what will be called when I create a new `Circle`.

To create a new Circle I call `Circle$new(<arguments passed to initialize>)`.
Below I create a `Circle` with radius 2 and print the radius and calculate the area.

```{r}
c1 <- Circle$new(radius = 2)
print(c(c1$radius, c1$area()))
```

I can change the radius by simply assigning it a new value.

```{r}
c1$radius <- 4
print(c(c1$radius, c1$area()))
```

You can also create private functions/variables.
For a more in-depth look at how to use `R6`,
check the link above or other sites.
This is not meant to be a comprehensive explanation.

## How to use active bindings

Suppose we wanted to add `diameter` to the `Circle` class.
I could add it as a function, just as I already have `area`.
I could even write the function to take a new value of diameter
so that it would update the radius at the same time.
But instead suppose I want to have it as a variable that can be set.
Then I'd be able to change the radius or diameter.
But then whenever I change one, I'd need to change the other
so they are in agreement.
Here's where I can use active bindings so it still looks like
they are variables despite them actually acting like functions.

Below I define the `Circle2` class to do this.
An active binding it actually a function.
If it is assigned a value, then the corresponding function
is evaluated, using the value given as the argument `value`.
If no value is given, i.e. the field is just called and not assigned,
then the `value` argument is missing.
The function should be defined to act one way when assigned
a value, and another when not.

```{r}

Circle2 <- R6::R6Class(
  classname = "Circle2",
  public = list(
    radius = NULL,
    area = function() {
      pi * self$radius ^ 2
    },
    initialize = function(radius) {
      self$radius <- radius
    }
  ),
  active = list(
    diameter = function(value) {
      if (missing(value)) {2*self$radius}
      else self$radius <- value / 2
    }
  )
)
```

Now if I create one of these objects, I can set the `radius`,
access the `diameter` as a field,
and then when I set the `diameter` it will automatically change
the `radius` to match it.

```{r}
c2 <- Circle2$new(radius=2)
print(c(c2$radius, c2$diameter, c2$area()))
c2$diameter <- 10
print(c(c2$radius, c2$diameter, c2$area()))
```

On my recent projects I've considered using active bindings
when I have a variable that I want to store in two formats,
e.g. as itself and on a log scale.
My main concern with this is that using active bindings will be slow.


## Are active bindings slow?

My goal of writing this is to find out how slow active bindings are to use.
It seems like there should be an overhead cost to use something
that is flexible and requires the machine determining which action to take.
I will use `microbenchmark` to time these and see how fast they are.

First I'll just try to access the `radius`, a normal field, from `c2`,
the `diameter`, an active binding, of `c2`,
and compare these two just multiplying the `radius` by 2.

```{r}
microbenchmark::microbenchmark(
  c2$radius,
  c2$diameter,
  {c2$radius * 2}
)
```

First we see by comparing the first and third line that the multiplication
by 2 takes almost no time.
Second by comparing these to the second line we see that there
is a small cost to using the active binding.
On my computer this is less than 2 microseconds,
meaning that it is negligible unless calling these tons of times.

Next I want to compare how much longer it takes to
assign an active binding.
Here we set the `radius` and `diameter` to a random value.
Recall that setting the active binding really is just
calling a function to set the radius to half of the given value.

```{r}
microbenchmark::microbenchmark(
  {c2$radius <- runif(1)},
  {c2$diameter <- runif(1)}
)
```

We see that it takes twice as long to set the active binding,
but the difference on my computer is still only 3 microseconds.
This tells me that I shouldn't worry about the slowdown of
active bindings unless I am calling them thousands or millions of times.
And since the code I was considering using them on did lots of
matrix equation solving, it wouldn't hurt the performance
in any noticeable way.


## Conclusion

Active bindings are a useful part of `R6` classes when the
fields are entangled in some way.
They can also be used when randomness is involved in each
access, which I did not cover here.
Accessing them is quite fast and should not cause any speed problems
unless they are called millions of times.




