---
title: Using crossprod in R
author: Collin Erickson
date: '2018-04-23'
slug: using-crossprod-in-r
categories: []
tags: []
---

Often I have come across the function `crossprod` in R, and found that it didn't have much effect.
Here I will explore if it is useful and when.


The function `crossprod(x,y)` is equivalent to `t(x) %*% y`, i.e.,
allows you to multiply two matrices or vectors, where the transpose
of the first argument is used.
There is also the function `tcrossprod` to transpose the second argument.
The function allows you to avoid the operation of finding the transpose,
and thus should be faster.

Disclaimer: this time speedup is tiny and should only be used if your code
is slow and you have identified the bottleneck.
99% of R users don't need to worry about minor time speedups
and can ignore this.

```{r}
set.seed(0)
x <- matrix(rnorm(20), 10, 2)
y <- matrix(rnorm(20), 10, 2)
```

```{r}
t(x) %*% y
```

```{r}
crossprod(x, y)
```

So we see that they give the same answer.

Let me try to benchmark it to see if there is a speedup.

```{r}
microbenchmark::microbenchmark(
  t(x) %*% y,
  crossprod(x, y)
)
```

I see about a 5 times speedup.
I'm surprised the difference is this big for such small matrices.

And for `tcrossprod`

```{r}
microbenchmark::microbenchmark(
  x %*% t(y),
  tcrossprod(x, y)
)
```
Still about a 3x speedup.

### Larger example

Let's try an example with larger matrices.

```{r}
set.seed(1)
x <- matrix(rnorm(1000), 100, 10)
y <- matrix(rnorm(1000), 100, 10)
```


```{r}
microbenchmark::microbenchmark(
  t(x) %*% y,
  crossprod(x, y)
)
```

Now the speedup is 50% reduction in speed.

```{r}
microbenchmark::microbenchmark(
  x %*% t(y),
  tcrossprod(x, y)
)
```

And there is almost no difference for the reverse.

### Even bigger example


```{r}
set.seed(2)
n <- 300
x <- matrix(rnorm(n^2), n, n)
y <- matrix(rnorm(n^2), n, n)
```


```{r}
microbenchmark::microbenchmark(
  t(x) %*% y,
  crossprod(x, y)
)
```

Now using `crossprod` takes twice as long.
I have no clue why it would take longer.
This is surprising.
I would think that the speedup would be larger with larger matrices
since it's harder to form the transpose.

```{r}
microbenchmark::microbenchmark(
  x %*% t(y),
  tcrossprod(x, y)
)
```

And they are about the same for the reverse.


## Conclusion

`crossprod` and `tcrossprod` are functions in R for multiplying
a transpose of a matrix and another matrix without explicitly
forming the transpose yourself first using `t`.
This should be faster since it saves a step.
With small matrices I saw a speedup of about 5x using `crossprod`.
With larger matrices there is almost no difference or a slowdown of 2x.

I'm just as confused about this function as when I started,
but now I know that it can be worth trying and benchmarking
to see if it helps.
However, it is not worth bothering with unless speed is essential
and you have identified the matrix multiplication as a bottleneck.
