---
title: Getting started with R package mlbgameday
author: Collin Erickson
date: '2019-03-18'
slug: getting-started-with-r-package-mlbgameday
categories: []
tags: [R, mlbgameday, baseball]
---

I'm going to start digging into baseball statistics.
There are a couple of R packages that facilitate the collecting of data
from MLB.
I'll be using the R package mlbgameday, mostly because I saw that it was
recently updated.

## Getting some data

First we load the package.
Along with a couple of other of packages that we'll need.

```{r}
library(mlbgameday)
library(dplyr)
library(ggplot2)
library(magrittr)
```

Now we'll grab some data.
I'm picking the date 8/22/18 for no specific reason except
that it was fairly recent.
This chunk can be a little slow since it has to get it from the web.

```{r}
dat <- get_payload(start = "2018-08-22", end = "2018-08-22")
```

```{r}
dat %>% str
```

The data returned is a whole mess of things.
But we can see that there are a few different main categories.

```{r}
dat %>% names
```

### What's in atbat?

```{r}
dat$atbat %>% str
```

Looks like a description of every at bat, unsurprisingly.

### action?

```{r}
dat$action %>% str
```

Looks like this has action that is not the result of the at bat,
such as a mound visit or pitching change.

### pitch?

```{r}
dat$pitch %>% str
```

This has the result of every pitch.
This is where most of the fun will be, such as looking at balls and strikes.

### runner?

```{r}
dat$runner %>% str
```

Looks like this shows the movement of runners on the bases.

### po?

```{r}
dat$po %>% str
```

Looks like this just has information about pickoff attempts,
not very exciting.


## Conclusion

This has been a very short dive into what kind of data you can get from
the R package mlbgameday.
I'm going to try to do some actual data analysis next
to see what I kind dig up.
