---
title: BABIP
author: Collin Erickson
date: '2019-05-07'
slug: babip
categories: []
tags: [R, mlbgameday, baseball]
---

Recently I was reading about pitchers having no control over results of a ball in play.
This is the main idea behind FIP/DIPS.
This goes against the common idea that pitchers can be good if they induce weak contact.
I'm going to investigate this in this post.

Similar to last time, I'm using the same data from `mlbgameday`.

```{r}
library(mlbgameday)
library(magrittr)
library(ggplot2)
library(dplyr)
```

Get data.

```{r, eval=F}
# Takes hours to get all the data for the year
dat <- get_payload(start = "2018-01-01", end = "2018-12-31")
```


```{r, include=F}
dat <- readRDS("C:/Users/cbe117/Documents/GitHub/website-hugo//scratch/MLB2018.rds")
dat$pitch$az <- as.numeric(dat$pitch$az)
dat$pitch$break_y <- as.numeric(dat$pitch$break_y)
dat$pitch$break_angle <- as.numeric(dat$pitch$break_angle)
```

Get pitcher names from at bat data.
`play_guid` is included since we
only want the final pitch of each at bat, not every pitch.
```{r}
d2 <- inner_join(dat$pitch, dat$atbat, by=c("num", "url", "play_guid"))
```

View all at bat outcomes.

```{r}
d2$event %>% table
```

The equation for BABIP is 

$BABIP = \frac{H - HR}{AB - K - HR - SF} $

Essentially we have non-homerun hits over non-homerun hits plus outs from balls in play.
We need to find these two categories from the data.

```{r}
inplayhits <- c("Double", "Single", "Triple")
inplayouts <- c("Bunt Groundout", "Bunt Lineout", "Bunt Pop Out", "Double Play", "Field Error", "Fielders Choice", "Fielders Choice Out", "Flyout", "Forceout", "Grounded Into DP", "Groundout", "Lineout", "Pop Out", "Triple Play")
```

To make sure I didn't miss any, I'll check the ones not in any of these.

```{r}
d2 %>% filter(!(event %in% c(inplayhits, inplayouts))) %>% .$event %>% table
```

Now we have our inplay data.

```{r}
inplay <- d2 %>% filter(event %in% c(inplayhits, inplayouts))
```

We can find the league-wide BABIP, which I think should be around .300.

```{r}
sum(inplay$event %in% inplayhits) / nrow(inplay)
```

Wow, that was really close and I hadn't checked it beforehand.

Now we want to see how many pitchers can control this value.

First we'll check the distribution over all pitchers above a certain number of in play at bats.

```{r}
d2 %>% group_by(pitcher_name) %>% summarize(ipo=sum(event %in% inplayouts), iph=sum(event %in% inplayhits), N=iph+ipo, BABIP=iph/(iph+ipo)) %>% filter(N>100) %>% arrange(BABIP)
```

We can see that the pitchers with lowest BABIP are mostly relievers,
either because they are better or small sample size.

In looking through the data, I found an error: spring training data has been included.

```{r}
d2 %>% filter(pitcher_name=="Rex Brothers") %>% .$date
```

I don't see an easy way to filter out spring training games.
(There is when collecting the data, but not after getting it.)
I'll just filter out all the dates up through March, which will cut off
a little of the regular season.

```{r}
d2 <- d2 %>% filter(as.character(date) >= "2018-04-01")
d2$date %>% unique
```

Now let's check the league BABIP again.

```{r}
inplay <- d2 %>% filter(event %in% c(inplayhits, inplayouts))
sum(inplay$event %in% inplayhits) / nrow(inplay)
```

And the pitchers with the best BABIP.
```{r}
d2 %>% group_by(pitcher_name) %>% summarize(ipo=sum(event %in% inplayouts), iph=sum(event %in% inplayhits), N=iph+ipo, BABIP=iph/(iph+ipo)) %>% filter(N>100) %>% arrange(BABIP)
```

I'm going to create a data frame for this.
20,000 pitches have `NA` for `pitcher_name`, so we'll remove those.
```{r}
BABIPbypitcher <- d2 %>% group_by(pitcher_name) %>% summarize(ipo=sum(event %in% inplayouts), iph=sum(event %in% inplayhits), N=iph+ipo, BABIP=iph/(iph+ipo)) %>% filter(N>100) %>% arrange(BABIP) %>% filter(!is.na(pitcher_name))
```

```{r}
ggplot(BABIPbypitcher, aes(N, BABIP)) + geom_point()
```

Now we have an idea of what this distribution looks like.
But we need a way to see how much control a pitcher has on this number.

Some ideas:

1. For each pitcher, split data randomly into two groups, compare BABIP in each group.

2. For each pitcher, split data in first/second half of season, compare BABIP in each group.

3. See if there's any correlation between pitch speed and BABIP. 
This would assume faster pitches are hard to hit.
Same argument could be made for pitches that break a lot.

4. Find quality of each pitch by seeing how often it gets a swinging strike vs a homerun,
then see if BABIP is related to the pitch quality.

5. Fit a logistic model to predict hit/out using pitch speed, location, break, to see if there
are types of pitches that lead to outs more often.

### 1. Randomly split at bats for each pitcher

I'll split each ball in play for each pitcher into two random groups.
Then we'll compare the BABIP in each group.

```{r}
d2 %>% mutate(group12=sample(1:2,n(), T)) %>% group_by(pitcher_name, group12) %>% summarize(ipo=sum(event %in% inplayouts), iph=sum(event %in% inplayhits), N=iph+ipo, BABIP=iph/(iph+ipo))  %>% filter(!is.na(pitcher_name)) %>% group_by(pitcher_name) %>% summarize(BABIP1=BABIP[1], BABIP2=BABIP[2], N1=N[1], N2=N[2], N=N1+N2) %>% filter(N1>80 & N2>80) %>% {ggplot(., aes(BABIP1, BABIP2)) + geom_point(aes(color=N1+N2)) + geom_smooth()}
```

There appears to be no correlation.
The correlation coefficient is about 0.04.

```{r}
d2 %>% mutate(group12=sample(1:2,n(), T)) %>% group_by(pitcher_name, group12) %>% summarize(ipo=sum(event %in% inplayouts), iph=sum(event %in% inplayhits), N=iph+ipo, BABIP=iph/(iph+ipo))  %>% filter(!is.na(pitcher_name)) %>% group_by(pitcher_name) %>% summarize(BABIP1=BABIP[1], BABIP2=BABIP[2], N1=N[1], N2=N[2], N=N1+N2) %>% filter(N1>80 & N2>80) %>% {cor(.$BABIP1, .$BABIP2)}
```

It would be nice to see what this would look like comparing another quantity,
such as percentage of balls put in play.

```{r}
d2 %>% mutate(group12=sample(1:2,n(), T)) %>% group_by(pitcher_name, group12) %>% summarize(inplay=sum(event %in% c(inplayhits, inplayouts)), notinplay=sum(!(event %in% c(inplayouts, inplayhits))), N=inplay+notinplay, InPlayPer=inplay/(inplay+notinplay))  %>% filter(!is.na(pitcher_name)) %>% group_by(pitcher_name) %>% summarize(InPlayPer1=InPlayPer[1], InPlayPer2=InPlayPer[2], N1=N[1], N2=N[2], N=N1+N2) %>% filter(N1>80 & N2>80) %>% {ggplot(., aes(InPlayPer1, InPlayPer2)) + geom_point(aes(color=N1+N2)) + geom_smooth() + geom_text(aes(label=ifelse(N1+N2>850, pitcher_name, '')))}
```

We see from the play that there is decent correlation between the two groups.
The correlation coefficient is .61, which is significant.

```{r}
d2 %>% mutate(group12=sample(1:2,n(), T)) %>% group_by(pitcher_name, group12) %>% summarize(inplay=sum(event %in% c(inplayhits, inplayouts)), notinplay=sum(!(event %in% c(inplayouts, inplayhits))), N=inplay+notinplay, InPlayPer=inplay/(inplay+notinplay))  %>% filter(!is.na(pitcher_name)) %>% group_by(pitcher_name) %>% summarize(InPlayPer1=InPlayPer[1], InPlayPer2=InPlayPer[2], N1=N[1], N2=N[2], N=N1+N2) %>% filter(N1>80 & N2>80) %>% {cor(.$InPlayPer1, .$InPlayPer2)}
```


### 2. Split up each pitchers results by the two halves of the season

I'll do the same as before, but now assign groups so the first half of the pitchers in play
balls are in one group, the rest in the other.

```{r}
d2 %>% group_by(pitcher_name) %>% mutate(group12=rep(c(1,2), c(floor(n()/2),n()-floor(n()/2)))) %>% ungroup %>% group_by(pitcher_name, group12) %>% summarize(ipo=sum(event %in% inplayouts), iph=sum(event %in% inplayhits), N=iph+ipo, BABIP=iph/(iph+ipo))  %>% filter(!is.na(pitcher_name)) %>% group_by(pitcher_name) %>% summarize(BABIP1=BABIP[1], BABIP2=BABIP[2], N1=N[1], N2=N[2], N=N1+N2) %>% filter(N1>80 & N2>80) %>% {ggplot(., aes(BABIP1, BABIP2)) + geom_point(aes(color=N1+N2)) + geom_smooth()}
```

Again, no clear pattern, and a correlation coefficient near zero.

```{r}
d2 %>% group_by(pitcher_name) %>% mutate(group12=rep(c(1,2), c(floor(n()/2),n()-floor(n()/2)))) %>% ungroup %>% group_by(pitcher_name, group12) %>% summarize(ipo=sum(event %in% inplayouts), iph=sum(event %in% inplayhits), N=iph+ipo, BABIP=iph/(iph+ipo))  %>% filter(!is.na(pitcher_name)) %>% group_by(pitcher_name) %>% summarize(BABIP1=BABIP[1], BABIP2=BABIP[2], N1=N[1], N2=N[2], N=N1+N2) %>% filter(N1>80 & N2>80) %>% with(cor(BABIP1, BABIP2))
```

Let's check proportion of balls put in play as before, but with this new grouping method.

```{r}
d2 %>% group_by(pitcher_name) %>% mutate(group12=rep(c(1,2), c(floor(n()/2),n()-floor(n()/2)))) %>% ungroup %>% group_by(pitcher_name, group12) %>% summarize(inplay=sum(event %in% c(inplayhits, inplayouts)), notinplay=sum(!(event %in% c(inplayouts, inplayhits))), N=inplay+notinplay, InPlayPer=inplay/(inplay+notinplay))  %>% filter(!is.na(pitcher_name)) %>% group_by(pitcher_name) %>% summarize(InPlayPer1=InPlayPer[1], InPlayPer2=InPlayPer[2], N1=N[1], N2=N[2], N=N1+N2) %>% filter(N1>80 & N2>80) %>% {ggplot(., aes(InPlayPer1, InPlayPer2)) + geom_point(aes(color=N1+N2)) + geom_smooth() + geom_text(aes(label=ifelse(N1+N2>850, pitcher_name, '')))}
```

Again, there is a pattern with .59 correlation coefficient.

```{r}
d2 %>%  group_by(pitcher_name) %>% mutate(group12=rep(c(1,2), c(floor(n()/2),n()-floor(n()/2)))) %>% ungroup %>% group_by(pitcher_name, group12) %>% summarize(inplay=sum(event %in% c(inplayhits, inplayouts)), notinplay=sum(!(event %in% c(inplayouts, inplayhits))), N=inplay+notinplay, InPlayPer=inplay/(inplay+notinplay))  %>% filter(!is.na(pitcher_name)) %>% group_by(pitcher_name) %>% summarize(InPlayPer1=InPlayPer[1], InPlayPer2=InPlayPer[2], N1=N[1], N2=N[2], N=N1+N2) %>% filter(N1>80 & N2>80) %>% {cor(.$InPlayPer1, .$InPlayPer2)}
```

I'm going to skip 3 and 4 and go to 5.
Those seem like too much work.

### 5. Logistic regression to predict if ball in play is hit or out

We're going to need to decide which features to include in the model.
I worked backwards and removed all the ones I don't want, so it's a mess of code below.
Basically I want to keep the location, speed, and spin data.

```{r}
d2 %>% select(-c(gameday_link.x, gameday_link.y, end_tfs_zulu, event_num.x, event_num.y, event_es, event_es, event2_es, date, event4, pitcher_name, batter_name, event3, event2, next_.y, inning.y, inning_side.y, inning.x, inning_side.x, away_team_runs, home_team_runs, url, des, des_es, atbat_des, atbat_des_es, p_throws, b_height, stand, start_tfs, start_tfs_zulu, pitcher, batter, play_guid, code, count, on_3b, on_2b, on_1b, num, next_.x, type_confidence, id, type, x, y, o, cc, mt, tfs, tfs_zulu, sz_top, sz_bot, x0, y0, z0, vx0, vy0, vz0, sv_id, b, s)) %>% filter(event %in% c(inplayhits, inplayouts)) %>%  mutate(iph=ifelse(event %in% inplayhits, 1, 0))  %>% with(iph) %>% table
```

```{r}
# data in play
dip <- d2 %>% select(-c(gameday_link.x, gameday_link.y, end_tfs_zulu, event_num.x, event_num.y, event_es, event_es, event2_es, date, event4, pitcher_name, batter_name, event3, event2, next_.y, inning.y, inning_side.y, inning.x, inning_side.x, away_team_runs, home_team_runs, url, des, des_es, atbat_des, atbat_des_es, p_throws, b_height, stand, start_tfs, start_tfs_zulu, pitcher, batter, play_guid, code, count, on_3b, on_2b, on_1b, num, next_.x, type_confidence, id, type, x, y, o, cc, mt, tfs, tfs_zulu, sz_top, sz_bot, x0, y0, z0, vx0, vy0, vz0, sv_id, b, s)) %>% filter(event %in% c(inplayhits, inplayouts)) %>%  mutate(iph=ifelse(event %in% inplayhits, 1, 0)) %>% select(-event)
```

I'm going to randomly assign the training and testing.

```{r}
train_inds <- sample(1:nrow(dip), floor(.7*nrow(dip)), T)
dip_train <- dip[train_inds,]
dip_test  <- dip[-train_inds,]
```


First I'll fit a simpler logistc regression model that only uses three of the factors.

```{r}
mod1 <- glm(iph ~ start_speed + spin_rate + break_length, "binomial", dip_train)
pred1train <- plogis(predict(mod1, dip_train))
qplot(pred1train, dip_train$iph+rnorm(nrow(dip_train), 0, .04), alpha=.0001)
pred1test <- plogis(predict(mod1, dip_test))
qplot(pred1test, dip_test$iph+rnorm(nrow(dip_test), 0, .02))
dip_train %>% mutate(p1 = pred1train, iphf=as.factor(iph)) %>% {ggplot(data=., mapping=aes(x=p1, group=iphf, fill=iphf)) + geom_histogram(alpha=.6)}
dip_test %>% mutate(p1 = pred1test, iphf=as.factor(iph)) %>% {ggplot(data=., mapping=aes(x=p1, group=iphf, fill=iphf)) + geom_histogram(alpha=.6)}
```

The plots all show that the model is not able to find any difference
between the two groups.
The first and third plots shows that it can't even find a difference on the training data,
which should be easier than testing data.
Let's try another model with more features.

```{r}
mod2 <- glm(iph ~ start_speed + spin_rate + break_length + pfx_x + pfx_z + px + pz + break_y + nasty, "binomial", dip_train)
pred1train <- plogis(predict(mod2, dip_train))
qplot(pred1train, dip_train$iph+rnorm(nrow(dip_train), 0, .04), alpha=.0001)
pred1test <- plogis(predict(mod2, dip_test))
qplot(pred1test, dip_test$iph+rnorm(nrow(dip_test), 0, .02))
dip_test %>% mutate(p1 = pred1test, iphf=as.factor(iph)) %>% {ggplot(data=., mapping=aes(x=p1, group=iphf, fill=iphf)) + geom_histogram(alpha=.6)}
```

With these additional features, it still can't really tell the difference between the two groups.

### BABIP by zone

I'm curious to see how the zone the pitch is in affects the BABIP.

```{r}
d2 %>% group_by(zone) %>% summarize(ipo=sum(event %in% inplayouts), iph=sum(event %in% inplayhits), N=iph+ipo, BABIP=iph/(iph+ipo)) %>% filter(N>100) %>% arrange(BABIP)
```

```{r}

zones <- data.frame(
  zone = (c(rep(1:9, each=4), rep(11:14, each=6))),
  x = c(1,2,2,1,2,3,3,2,3,4,4,3,1,2,2,1,2,3,3,2,3,4,4,3,1,2,2,1,2,3,3,2,3,4,4,3,  0,1,1,2.5,2.5,0, 2.5,4,4,5,5,2.5, 2.5,5,5,4,4,2.5, 0,2.5,2.5,1,1,0),
  y = c(3,3,4,4,3,3,4,4,3,3,4,4,2,2,3,3,2,2,3,3,2,2,3,3,1,1,2,2,1,1,2,2,1,1,2,2,  2.5,2.5,4,4,5,5, 4,4,2.5,2.5,5,5, 0,0,2.5,2.5,1,1, 0,0,1,1,2.5,2.5)
)
zones$bs <- ifelse(as.numeric(zones$zone) <=9, "S","B")
```


```{r}
babipzonesdf <- d2 %>% group_by(zone) %>% summarize(ipo=sum(event %in% inplayouts), iph=sum(event %in% inplayhits), N=iph+ipo, BABIP=iph/(iph+ipo)) %>% filter(N>100) %>% arrange(BABIP) %>% inner_join(zones, by="zone")
ggplot(babipzonesdf) + geom_polygon(aes(x=x,y=y,fill=BABIP, group=zone), alpha=.8)  + scale_fill_gradientn(colours=c('blue','white','red'))
```

Here we can get something.
The BABIP is highest in the lower two thirds of the strike zone,
and lowest outside the strike zone.
This isn't too surprising since it's hard to get contact on pitches outside the zone.
But at the same time, this would mean that pitchers can control their BABIP
by pitching to specific zones.

### BABIP by nasty

Looks like nastiness is not a good predictor of BABIP by itself.

```{r}
dip %>% {ggplot(data=., mapping=aes(x=nasty, group=iph, fill=iph)) + geom_histogram(alpha=.6)}
```



### Conclusion

From some basic data exploration here,
I've become more convinced in the assumption behind FIP/DIPS:
that pitchers can't control the proportion of balls in play that become hits.
This means that saying that some pitchers are able to induce weak contact
is not really a thing.
I could do some exploration on this.
The only thing useful I found was that BABIP is lower outside of the strike zone,
which makes sense.
