---
title: Swings and Misses
author: Collin Erickson
date: '2019-05-06'
slug: swings-and-misses
categories: []
tags: [R, mlbgameday, baseball]
---

I going to work more with data from `mlbgameday`.


```{r}
library(mlbgameday)
library(magrittr)
library(ggplot2)
library(dplyr)
```

Get data.

```{r, eval=F}
# Takes hours to get all the data for the year
dat <- get_payload(start = "2018-01-01", end = "2018-12-31")
```


```{r, include=F}
dat <- readRDS("C:/Users/cbe117/Documents/GitHub/website-hugo//scratch/MLB2018.rds")
```

Get pitcher names from at bat data.
```{r}
d2 <- inner_join(dat$pitch, dat$atbat, by=c("num", "url"))
```

Pick out the swinging strikes.

```{r}
d2 %>% filter(des %in% c("Swinging Strike", "Swinging Strike (Blocked)")) %>% group_by(pitcher_name) %>% summarize(N=n()) %>% arrange(desc(N))

```

These names are the ones we'd expect.

Who has the highest proportion of pitches as swinging strikes?


```{r}
d2 %>% filter(!is.na(pitcher_name)) %>% group_by(pitcher_name) %>% summarize(N=n(), Nss = sum(des %in% c("Swinging Strike", "Swinging Strike (Blocked)"))) %>% mutate(pSS = Nss / N) %>% arrange(desc(pSS))
```

Unsurprisingly these are only guys with few pitches. 
Lets only pick out pitchers with at least 300 pitches.

```{r}
d2 %>% filter(!is.na(pitcher_name)) %>% group_by(pitcher_name) %>% summarize(N=n(), Nss = sum(des %in% c("Swinging Strike", "Swinging Strike (Blocked)"))) %>% mutate(pSS = Nss / N) %>% arrange(desc(pSS)) %>% filter(N >= 300)
```

The pitchers best at missing bats are Josh Hader, Hector Neris, and Marcus Walden.
Unsurprisingly these are all relievers.

If we set a higher bar for N, we can see the best starting pitchers.

```{r}
d2 %>% filter(!is.na(pitcher_name)) %>% group_by(pitcher_name) %>% summarize(N=n(), Nss = sum(des %in% c("Swinging Strike", "Swinging Strike (Blocked)"))) %>% mutate(pSS = Nss / N) %>% arrange(desc(pSS)) %>% filter(N >= 2000)
```


Let's plot SS rate by pitcher and plot N vs Nss.
This is a nasty plot. I put both axes on logs, used colors to show the best/worse pSS, and tried to add names.

```{r}
d2 %>% filter(!is.na(pitcher_name)) %>% group_by(pitcher_name) %>% summarize(N=n(), Nss = sum(des %in% c("Swinging Strike", "Swinging Strike (Blocked)"))) %>% mutate(pSS = Nss / N) %>% arrange(desc(pSS)) %>% filter(Nss>0, N>1000) %>% 
  {(ggplot(., mapping=aes(N, Nss, color=pSS)) + geom_point() + geom_text(aes(label=ifelse(pSS>.2 | pSS<.07 | Nss>500 | N > 4e3, pitcher_name, ''))) + scale_color_gradientn(colors=c("blue", "black", "red")) )}
```


### Where do swings and misses happen?


```{r}
d2 %>% group_by(zone) %>% summarize(N=n(), Nss = sum(des %in% c("Swinging Strike", "Swinging Strike (Blocked)"))) %>% mutate(pSS = Nss / N) %>% arrange(desc(pSS))
```

We have a lot of pitches with zone NA, so we'll need to exclude these.
I'm going to use the zones I created last time to plot these.

```{r}
zone_values <- data.frame()
zones <- data.frame(
  zone = (c(rep(1:9, each=4), rep(11:14, each=6))),
  x = c(1,2,2,1,2,3,3,2,3,4,4,3,1,2,2,1,2,3,3,2,3,4,4,3,1,2,2,1,2,3,3,2,3,4,4,3,  0,1,1,2.5,2.5,0, 2.5,4,4,5,5,2.5, 2.5,5,5,4,4,2.5, 0,2.5,2.5,1,1,0),
  y = c(3,3,4,4,3,3,4,4,3,3,4,4,2,2,3,3,2,2,3,3,2,2,3,3,1,1,2,2,1,1,2,2,1,1,2,2,  2.5,2.5,4,4,5,5, 4,4,2.5,2.5,5,5, 0,0,2.5,2.5,1,1, 0,0,1,1,2.5,2.5)
)
zones$bs <- ifelse(as.numeric(zones$zone) <=9, "S","B")
ggplot() + geom_polygon(aes(x=x,y=y,fill=as.factor(bs), group=zone), zones, alpha=.5)
```

Now we can plot the swinging strike rates by zone.

```{r}
SSdf <- d2 %>% group_by(zone) %>% summarize(N=n(), Nss = sum(des %in% c("Swinging Strike", "Swinging Strike (Blocked)"))) %>% mutate(pSS = Nss / N) %>% arrange(desc(pSS)) %>% filter(!is.na(zone)) %>% inner_join(zones)
ggplot(SSdf) + geom_polygon(aes(x=x,y=y,fill=pSS, group=zone), alpha=.8)  + scale_fill_gradientn(colours=c('blue','white','red'))
```

Again the results aren't too surprising.
Swings and misses occur most on low pitches outside the strike zone.
If they were in the strike zone, they would be more likely to make contact.
The lowest swinging strike rate is right in the center of the strike zone,
which should be the easiest place to make contact.

So far we've been looking at the rate of swings and misses as a proportion of all pitches.
What if we just look at pitches the batters swung at?
We'll have swings and misses on one hand, and contact on the other.
In other words, given that the batter swung, what is the proportion that were misses?

```{r}
SSdf2 <- d2 %>% group_by(zone) %>% summarize(N=n(), Ncontact=sum(des %in% c("Foul", "Foul (Runner Going)", "Foul Bunt", "Foul Tip", "In play, no out", "In play, out(s)", "In play, run(s)")), Nss = sum(des %in% c("Swinging Strike", "Swinging Strike (Blocked)"))) %>% mutate(pSSwung = Nss / (Nss + Ncontact)) %>% filter(!is.na(zone)) %>% inner_join(zones, by="zone")
ggplot(SSdf2) + geom_polygon(aes(x=x,y=y,fill=pSSwung, group=zone), alpha=.8)  + scale_fill_gradientn(colours=c('blue','white','red'))
```
It's highest below the strike zone, lowest in the center row of the strike zone.
Similar trend as when looking at swinging strikes as a proportion of all pitches.
