---
title: Calculating park factors
author: Collin Erickson
date: '2019-06-04'
slug: calculating-park-factors
categories: []
tags: [R, baseball]
---

I've been thinking a bit about park factors recently.
A lot of stats are pretty meaningless without context.
40 home runs for a Rockies player is not as good as 40
home runs for a Dodgers player because Coors is much more
hitter-friendly.
Park factors are used to determine how many runs are scored at
any given stadium compared to the rest of the league.
+/- stats like OPS+ and ERA- adjust for the stadium,
making comparisons for players across teams more fair.


I'm just going to see if I can do park factor calculations simply.
I'll start by just finding the ratio of runs scored at home divided by
runs scored away for each team.


```{r}
library(magrittr)
library(ggplot2)
library(dplyr)
```


Load the data as `d`.

```{r, include=F}
d <- readRDS("C:/Users/cbe117/Documents/GitHub/website-hugo//scratch/BaseballSavantData/MLB2018.rds")
```

```{r, eval=F}
d <- readRDS("MLB2018.rds")
```

```{r}
game_scores <- d %>% group_by(home_team, away_team, game_pk) %>%
  summarise(home_score=max(home_score), away_score=max(away_score))
```

```{r}
team_scores <- with(game_scores,
                   rbind(data.frame(team=home_team, score=home_score, home=1),
                         data.frame(team=away_team, score=away_score, home=-1))
)
```


```{r}
parkfactors <- team_scores %>% group_by(team, home) %>% summarize(totalruns=sum(score)) %>% 
  group_by(team) %>% summarize(ParkFactor=totalruns[which(home>.5)] / totalruns[which(home<(-.5))]) %>% 
  arrange(desc(ParkFactor))
parkfactors %>% print(n=30)
```

These park factors aren't correct, but they are neighborhood estimates.

## Using runs allowed

I didn't use runs scored by opponents in the above calculation.
By using the total runs scored in home games divided by
total runs scored in away games, it should reduce the
error by a bit.

```{r}
team_scores2 <- with(game_scores,
                   rbind(data.frame(team=home_team, score=home_score+away_score, home=1),
                         data.frame(team=away_team, score=away_score+home_score, home=-1))
)
```


```{r}
parkfactors2 <- team_scores2 %>% group_by(team, home) %>% summarize(totalruns=sum(score)) %>% 
  group_by(team) %>% summarize(ParkFactor=totalruns[which(home>.5)] / totalruns[which(home<(-.5))]) %>% 
  arrange(desc(ParkFactor))
parkfactors2 %>% print(n=30)
```

These are decent results, and are basically what [ESPN gives
for 2018 park factors](http://www.espn.com/mlb/stats/parkfactor/_/year/2018).

## Factors not considered

There are still a ton of things that affect the park factor
that this simple calculation doesn't consider.

* Quality of teams/pitchers. If a team plays more bad teams at home
than away, it could affect the result.
In general this should average out.

* Number of outs. I just treated all games equally,
but games that go to extra innings, or if the bottom of the ninth isn't played,
should be accounted for.

* Not only did I not account for the number of outs,
I didn't account for the number of home and away games.
Usually they are equal, but not with game 163s, or rain outs
that never get made up. This would be a simple fix.

* Interleague games. Having a DH in home games but not all away 
(or vice versa for NL teams) boosts the park factor 
(decrease for NL).

* Park factors may be different for right/left handed hitters

* I only used the data for a single year, but I could reduce
the uncertainty by using data from multiple years.

* There are different factors for different events.
Each park could have a factor for home runs, and a different one
for doubles.


## Best formula

I tried to find what the best formula is for calculating park factors.
The calculations become very difficult, despite seeming simple
from the calculations I made.
[Baseball Reference gives details on how to calculate Park Adjustments](https://www.baseball-reference.com/about/parkadjust.shtml),
and it's not easy to follow.


