---
title: Using mlbgameday to see what pitches get hit
author: Collin Erickson
date: '2019-04-29'
slug: using-mlbgameday-to-see-what-pitches-get-hit
categories: []
tags: [R, mlbgameday, baseball]
---

Load packages


```{r}
library(mlbgameday)
library(dplyr)
library(ggplot2)
library(magrittr)
```

Get data. Using arbitrary dates.


```{r}
dat <- get_payload(start = "2018-08-01", end = "2018-08-31")
```


[This site](https://fastballs.wordpress.com/2007/08/02/glossary-of-the-gameday-pitch-fields/)
is useful for deciphering fields.


`dat$pitch` has pitch data (location, speed, etc), `dat$atbat` has at bat data (result of at bat, etc),
so we need to join these.

```{r}
# d2 <- inner_join(dat$pitch, dat$atbat, by=c("num", "url")) # All pitches are kept
# Keep only final pitches of each at bat
d2 <- inner_join(dat$pitch, dat$atbat, by=c("play_guid", "num", "url"))
```

event has the result of the atbat

```{r}
d2$event %>% table %>% sort(decreasing = T)
```

Let's only keep the main events.

```{r}
d3 <- d2 %>% filter(event %in% c("Strikeout", "Groundout", "Single", "Flyout", "Walk", "Lineout", "Pop Out", "Double", "Home Run"))
```

There are 65 of these that don't have px or pz. I'll remove them.

```{r}
d3$px %>% summary()
d3 <- d3 %>% filter(!is.na(px))
```



Let's look at where these occur in the strike zone.

```{r}
ggplot(data=d3, mapping=aes(px, pz, color=event)) + geom_point(alpha=.2) + facet_wrap(. ~ event)
```

Unsurprisingly, walks are from out of the zone and extra base hits are in the zone.

```{r}
ggplot(data=d3, mapping=aes(px, pz, color=event)) + geom_point(alpha=.2) + facet_wrap(. ~ event) + coord_fixed() + stat_density_2d()
```

The zone will show which zone the pitched ended up in. 3x3 for the strike zone, four ball zones.
These may change across batters (there is overlap comparing 3v4).

```{r}
ggplot(data=d3, mapping=aes(px, pz, color=event)) + geom_point(alpha=.04) + facet_wrap(. ~ zone) + coord_fixed()
```

Here I'm making zones for the sake of plotting heat maps.

```{r}

x <- c()
zone_values <- data.frame()
zones <- data.frame(
  zone = (c(rep(1:9, each=4), rep(11:14, each=6))),
  x = c(1,2,2,1,2,3,3,2,3,4,4,3,1,2,2,1,2,3,3,2,3,4,4,3,1,2,2,1,2,3,3,2,3,4,4,3,  0,1,1,2.5,2.5,0, 2.5,4,4,5,5,2.5, 2.5,5,5,4,4,2.5, 0,2.5,2.5,1,1,0),
  y = c(3,3,4,4,3,3,4,4,3,3,4,4,2,2,3,3,2,2,3,3,2,2,3,3,1,1,2,2,1,1,2,2,1,1,2,2,  2.5,2.5,4,4,5,5, 4,4,2.5,2.5,5,5, 0,0,2.5,2.5,1,1, 0,0,1,1,2.5,2.5)
)
zones$bs <- ifelse(as.numeric(zones$zone) <=9, "S","B")
ggplot() + geom_polygon(aes(x=x,y=y,fill=as.factor(bs), group=zone), zones, alpha=.5)
```

Okay, so now I want to show that for a given zone, what is the probability of it being a home run (or other event)?

First we'll just group the events by their zone.
```{r}
zone_events <- d3 %>% group_by(event, zone) %>% summarize(n=n())
zone_events %>% head
```

```{r}
zone_events_total <- d3 %>% group_by(zone) %>% summarize(ntotal=n())
zone_events_total
```

Now we join the zone_events with the total

```{r}
zone_events2 <- dplyr::full_join(zone_events, zone_events_total, by="zone") %>% mutate(p=n/ntotal)
zone_events2 %>% head
```

Now I'll join with the zone locations.

```{r}
zone_events3 <- full_join(zone_events2, zones, by="zone")
zone_events3 %>% head
```

Now let's try to plot it.

```{r}
ggplot(zone_events3) + geom_polygon(aes(x=x,y=y,fill=p, group=zone), alpha=.8)  + scale_fill_gradientn(colours=c('blue','white','red')) + facet_wrap(. ~ event)
```

This plot looks reasonable.
It says pitches in the bottom left lead to strikeouts.
Pitches in the strike zone don't lead to walks, and pitches in the strike zone don't lead to walks.
Groundouts result more often form pitches down than up.

It doesn't look that good or useful.
But it's a start.
