---
title: Looking at pitches with the R package mlbgameday
author: Collin Erickson
date: '2019-03-19'
slug: looking-at-pitches-with-the-r-package-mlbgameday
categories: []
tags: [R, mlbgameday, baseball]
---

I'll be continuing on from the last post.

Load packages first.

```{r}
library(mlbgameday)
library(dplyr)
library(ggplot2)
library(magrittr)
```

Get some data.


```{r}
dat <- get_payload(start = "2018-08-22", end = "2018-08-22")
```

We're going to be looking at dat$pitch.

```{r}
dat$pitch %>% str
```

It looks like x and y are the coordinates where the pitch crosses the plate.
And `type` is probably ball/strike.

```{r}
ggplot(data=dat$pitch, mapping=aes(x, y, color=type)) + geom_point()
```

Let's try faceting this into three plots.

```{r}
ggplot(data=dat$pitch, mapping=aes(x, y, color=type)) + geom_point() +facet_grid(. ~ type)
```

It looks like B is ball, but it's not clear exactly what
the difference is between S and X.
Probably X is balls put in play, but does that include fouls?
We'll also want to separate out called strikes from swinging strikes.
It looks like `des` gives a further distinction.

```{r}
paste(dat$pitch$des, "///", dat$pitch$type) %>% table
```

Indeed B is ball, but it also includes hit by pitch, pitchout,
and automatic ball (not sure what that is).
S includes strikes and fouls.
X exclusively indicates balls put in play.

I'm going to filter out some of these des values that don't occur 
very often using functions from dplyr.

```{r}
pdf <- dat$pitch %>% group_by(des) %>% filter(n() > 100) %>% ungroup
```

```{r}
ggplot(data=pdf, mapping=aes(x, y, color=des)) + geom_point()
```


```{r}
ggplot(data=pdf, mapping=aes(x, y, color=des)) + geom_point() + facet_wrap(. ~ des)
```

Nothing too surprising here.
Now we can see that the division between Ball and Called Strike
can be used to estimate the strike zone.
I'm going to filter so only those two are left.

```{r}
pdf2 <- (pdf %>% filter(des=="Ball" | des=="Called Strike"))
```


```{r}
ggplot(data=pdf2, mapping=aes(x, y, color=des)) + geom_point(alpha=.25) + facet_wrap(. ~ des)
```


## Using machine learning to classify balls and strikes

Now we can use this data to create a model that will
be able to predict balls and strikes.

I'll create a grid of data points that we can predict on and plot.

```{r}
testdf <- expand.grid(x=seq.int(50,200,1), y=seq.int(100,250,1))
```


### k nearest neighbors

The first method I'll try is k nearest neighbors since it
seems like it should work easily and it's easy to understand.
To predict whether a given pitch is a ball or strike,
we simply find the k nearest neighbors (shortest distance
from the point), and assign it the class that is the majority
of the classes of those k points.

First we'll use $k=1$.

```{r, eval=F}
knn1pred <- rep(NA, nrow(testdf))
apply(testdf, 1, 
      function(rowi) {
        print(rowi)
        dists <- sapply(1:nrow(pdf2), function(gi) {sum((rowi - pdf2[gi,c('x','y')])^2)})
        inc <- which.min(dists)[1]
        pdf2$des[inc]
      }
)
```

This is way too slow, and probably not even correct.
I had to kill it because it would take all day to finish.
I tried to do a simple implementation, but it is not efficient enough.

It's always better to try to use an already existing package
that has been optimized for performance.
In this case we can use the R package `class`, and it's function
`knn`.

```{r}
knn1_pred <- class::knn(train = pdf2[,c('x','y')], test = testdf, k = 1, pdf2$des)
```

Now let's plot it and look at the classifications

```{r}
knn1_pred2 <- data.frame(x=testdf$x,
                  y=testdf$y,
                  class=knn1_pred)
ggplot(data=knn1_pred2, mapping=aes(x, y, color=class)) + geom_point() + facet_grid(. ~ class)
```

This looks reasonable based on the data we have.
It not very smooth at the edges.
We can try using a larger value for $k$.
This will help smooth it out, but we need to be careful
that it doesn't get smoothed too much.
Let's try using $k=11$.

```{r}
knn11_pred <- class::knn(train = pdf2[,c('x','y')], test = testdf, k = 11, pdf2$des)
```


```{r}
knn11_pred2 <- data.frame(x=testdf$x,
                  y=testdf$y,
                  class=knn11_pred)
ggplot(data=knn11_pred2, mapping=aes(x, y, color=class)) + geom_point() + facet_grid(. ~ class)
```

It is definitely smoother.
We might hope for a perfect rectangle, but that would be
unrealistic because umpires don't make perfect calls,
and we are limited by the data that we have.

### Random forest

```{r}
rf_pred <- randomForest::randomForest(x = as.matrix(pdf2[,c('x','y')]), xtest = as.matrix(testdf), y=as.factor(pdf2$des))
```

Now let's plot it and look at the classifications

```{r}
rf_pred <- data.frame(x=testdf$x,
                  y=testdf$y,
                  class=rf_pred$test$predicted)
ggplot(data=rf_pred, mapping=aes(x, y, color=class)) + geom_point() + facet_grid(. ~ class)
```

Again, this looks reasonable.



## Conclusion

We've taken MLB data and created a model that can predict whether
a given pitch should be classified as a ball or strike based on
its coordinates when it crossed the plate.

With more data, we can look into how the strike zone changes
based on the handedness of the batter,
the type of pitch,
and the count.
