---
title: Getting started with RStan
author: Collin Erickson
date: '2019-06-02'
slug: getting-started-with-rstan
categories: []
tags: [RStan, R]
---

I'm trying to learn more about Bayesian modeling,
and I want to try to use Bugs/Jags/RStan.
RStudio seems to have the most support for RStan
(File -> New file has an option for a Stan file,
but not the others).
So my goal for now is to get a simple example working
using RStan.

I started by following this blog post:
https://www.r-bloggers.com/hierarchical-models-with-rstan-part-1/

## Getting errors

Here's the data in R that needs to be included.

```{r}
#load libraries
library(rstan)
library(RColorBrewer)
#simulate some data
set.seed(20161110)
N<-100 #sample size
J<-10 #number of plant species
id<-rep(1:J,each=10) #index of plant species
K<-3 #number of regression coefficients
#population-level regression coefficient
gamma<-c(2,-1,3)
#standard deviation of the group-level coefficient
tau<-c(0.3,2,1)
#standard deviation of individual observations
sigma<-1
#group-level regression coefficients
beta<-mapply(function(g,t) rnorm(J,g,t),g=gamma,t=tau) 
#the model matrix
X<-model.matrix(~x+y,data=data.frame(x=runif(N,-2,2),y=runif(N,-2,2)))
y<-vector(length = N)
for(n in 1:N){
  #simulate response data
  y[n]<-rnorm(1,X[n,]%*%beta[id[n],],sigma)
}
```

```{r}
rstan_options(auto_write = TRUE)
```


Then I copied the Stan model into a file `stan-test2.stan`.
I used the second model  with a non-centered parameterization.
Then when I tried to run it, I got an error.

```{r, eval=F}
#run the model
m_hier<-stan(file="./stan-test2.stan",data=list(N=N,J=J,K=K,id=id,X=X,y=y))
```

```{r, eval=F, include=F}
#run the model
m_hier<-stan(file="./scratch/stan-test2.stan",data=list(N=N,J=J,K=K,id=id,X=X,y=y))
```

```
Error in compileCode(f, code, language = language, verbose = verbose) : 
  Compilation ERROR, function(s)/method(s) not created! Error in .shlib_internal(commandArgs(TRUE)) : 
  C++14 standard requested but CXX14 is not defined
Calls: <Anonymous> -> .shlib_internal
Execution halted
In addition: Warning messages:
1: In readLines(file, warn = TRUE) :
  incomplete final line found on 'C:\Users\cbe117\Documents\GitHub\website-hugo\scratch\stan-test2.stan'
2: In system(cmd, intern = !verbose) :
  running command 'C:/PROGRA~1/R/R-36~1.0/bin/x64/R CMD SHLIB file8403fc52d89.cpp 2> file8403fc52d89.cpp.err.txt' had status 1
Error in sink(type = "output") : invalid connection
```

After some Googling, I found a suggestion to add a line with 
`CXX14 = g++ -std=c++1y -Wno-unused-variable -Wno-unused-function -fPIC`
into my Makevars.win file.
This removed the error, but I had another one.


```{r, eval=F}
#run the model
m_hier<-stan(file="./stan-test2.stan",data=list(N=N,J=J,K=K,id=id,X=X,y=y))
```

```
Error in compileCode(f, code, language = language, verbose = verbose) : 
  Compilation ERROR, function(s)/method(s) not created! file84038d44756.cpp:1:0: warning: -fPIC ignored for target (all code is position independent)
 
 ^
In file included from C:/Users/cbe117/Documents/R/win-library/3.6/BH/include/boost/random/detail/integer_log2.hpp:19:0,
                 from C:/Users/cbe117/Documents/R/win-library/3.6/BH/include/boost/random/detail/int_float_pair.hpp:26,
                 from C:/Users/cbe117/Documents/R/win-library/3.6/BH/include/boost/random/exponential_distribution.hpp:27,
                 from C:/Users/cbe117/Documents/R/win-library/3.6/BH/include/boost/random/gamma_distribution.hpp:25,
                 from C:/Users/cbe117/Documents/R/win-library/3.6/StanHeaders/include/stan/math/prim/mat/prob/dirichlet_rng.hpp:5,
                 from C:/Users/cbe117/Documents/R/win-library/3.6/StanHeaders/include/stan/math/prim/mat.hpp:276,
                 from C:/Users/cbe117/Documents/R/win-library/3.6/StanHeaders/include/stan/math/rev/mat.hpp:12,
Error in sink(type = "output") : invalid connection
```

I was able to fix this by following the suggestion from [this StackOverflow answer](
https://stackoverflow.com/questions/56256884/error-on-the-stan-file-compilation-using-r-3-6-0-and-win-10)
I added a line with `CXX14 = "C:\Rtools\mingw_64\bin\g++.exe"` to my Makevars.win file,
and then it ran fine.


## Running the model

Now the model works and gives the following output.

```{r}
m_hier<-stan(file="./../../content/post/stan-test2.stan",data=list(N=N,J=J,K=K,id=id,X=X,y=y))
```

```
SAMPLING FOR MODEL 'stan-test2' NOW (CHAIN 1).
Chain 1: 
Chain 1: Gradient evaluation took 0.001 seconds
Chain 1: 1000 transitions using 10 leapfrog steps per transition would take 10 seconds.
Chain 1: Adjust your expectations accordingly!
Chain 1: 
Chain 1: 
Chain 1: Iteration:    1 / 2000 [  0%]  (Warmup)
Chain 1: Iteration:  200 / 2000 [ 10%]  (Warmup)
Chain 1: Iteration:  400 / 2000 [ 20%]  (Warmup)
Chain 1: Iteration:  600 / 2000 [ 30%]  (Warmup)
Chain 1: Iteration:  800 / 2000 [ 40%]  (Warmup)
Chain 1: Iteration: 1000 / 2000 [ 50%]  (Warmup)
Chain 1: Iteration: 1001 / 2000 [ 50%]  (Sampling)
Chain 1: Iteration: 1200 / 2000 [ 60%]  (Sampling)
Chain 1: Iteration: 1400 / 2000 [ 70%]  (Sampling)
Chain 1: Iteration: 1600 / 2000 [ 80%]  (Sampling)
Chain 1: Iteration: 1800 / 2000 [ 90%]  (Sampling)
Chain 1: Iteration: 2000 / 2000 [100%]  (Sampling)
Chain 1: 
Chain 1:  Elapsed Time: 1.112 seconds (Warm-up)
Chain 1:                1.288 seconds (Sampling)
Chain 1:                2.4 seconds (Total)
Chain 1: 

SAMPLING FOR MODEL 'stan-test2' NOW (CHAIN 2).
Chain 2: Rejecting initial value:
Chain 2:   Error evaluating the log probability at the initial value.
Chain 2: Exception: normal_lpdf: Scale parameter is -1.97613, but must be > 0!  (in 'model84027fb29ff_stan_test2' at line 38)

Chain 2: 
Chain 2: Gradient evaluation took 0 seconds
Chain 2: 1000 transitions using 10 leapfrog steps per transition would take 0 seconds.
Chain 2: Adjust your expectations accordingly!
Chain 2: 
Chain 2: 
Chain 2: Iteration:    1 / 2000 [  0%]  (Warmup)
Chain 2: Iteration:  200 / 2000 [ 10%]  (Warmup)
Chain 2: Iteration:  400 / 2000 [ 20%]  (Warmup)
Chain 2: Iteration:  600 / 2000 [ 30%]  (Warmup)
Chain 2: Iteration:  800 / 2000 [ 40%]  (Warmup)
Chain 2: Iteration: 1000 / 2000 [ 50%]  (Warmup)
Chain 2: Iteration: 1001 / 2000 [ 50%]  (Sampling)
Chain 2: Iteration: 1200 / 2000 [ 60%]  (Sampling)
Chain 2: Iteration: 1400 / 2000 [ 70%]  (Sampling)
Chain 2: Iteration: 1600 / 2000 [ 80%]  (Sampling)
Chain 2: Iteration: 1800 / 2000 [ 90%]  (Sampling)
Chain 2: Iteration: 2000 / 2000 [100%]  (Sampling)
Chain 2: 
Chain 2:  Elapsed Time: 1.237 seconds (Warm-up)
Chain 2:                1.184 seconds (Sampling)
Chain 2:                2.421 seconds (Total)
Chain 2: 

SAMPLING FOR MODEL 'stan-test2' NOW (CHAIN 3).
Chain 3: Rejecting initial value:
Chain 3:   Error evaluating the log probability at the initial value.
Chain 3: Exception: normal_lpdf: Scale parameter is -1.05795, but must be > 0!  (in 'model84027fb29ff_stan_test2' at line 38)

Chain 3: Rejecting initial value:
Chain 3:   Error evaluating the log probability at the initial value.
Chain 3: Exception: normal_lpdf: Scale parameter is -0.667925, but must be > 0!  (in 'model84027fb29ff_stan_test2' at line 38)

Chain 3: 
Chain 3: Gradient evaluation took 0 seconds
Chain 3: 1000 transitions using 10 leapfrog steps per transition would take 0 seconds.
Chain 3: Adjust your expectations accordingly!
Chain 3: 
Chain 3: 
Chain 3: Iteration:    1 / 2000 [  0%]  (Warmup)
Chain 3: Iteration:  200 / 2000 [ 10%]  (Warmup)
Chain 3: Iteration:  400 / 2000 [ 20%]  (Warmup)
Chain 3: Iteration:  600 / 2000 [ 30%]  (Warmup)
Chain 3: Iteration:  800 / 2000 [ 40%]  (Warmup)
Chain 3: Iteration: 1000 / 2000 [ 50%]  (Warmup)
Chain 3: Iteration: 1001 / 2000 [ 50%]  (Sampling)
Chain 3: Iteration: 1200 / 2000 [ 60%]  (Sampling)
Chain 3: Iteration: 1400 / 2000 [ 70%]  (Sampling)
Chain 3: Iteration: 1600 / 2000 [ 80%]  (Sampling)
Chain 3: Iteration: 1800 / 2000 [ 90%]  (Sampling)
Chain 3: Iteration: 2000 / 2000 [100%]  (Sampling)
Chain 3: 
Chain 3:  Elapsed Time: 1.317 seconds (Warm-up)
Chain 3:                0.982 seconds (Sampling)
Chain 3:                2.299 seconds (Total)
Chain 3: 

SAMPLING FOR MODEL 'stan-test2' NOW (CHAIN 4).
Chain 4: Rejecting initial value:
Chain 4:   Error evaluating the log probability at the initial value.
Chain 4: Exception: normal_lpdf: Scale parameter is -0.704333, but must be > 0!  (in 'model84027fb29ff_stan_test2' at line 38)

Chain 4: Rejecting initial value:
Chain 4:   Error evaluating the log probability at the initial value.
Chain 4: Exception: normal_lpdf: Scale parameter is -0.366959, but must be > 0!  (in 'model84027fb29ff_stan_test2' at line 38)

Chain 4: Rejecting initial value:
Chain 4:   Error evaluating the log probability at the initial value.
Chain 4: Exception: normal_lpdf: Scale parameter is -1.53722, but must be > 0!  (in 'model84027fb29ff_stan_test2' at line 38)

Chain 4: Rejecting initial value:
Chain 4:   Error evaluating the log probability at the initial value.
Chain 4: Exception: normal_lpdf: Scale parameter is -1.56318, but must be > 0!  (in 'model84027fb29ff_stan_test2' at line 38)

Chain 4: Rejecting initial value:
Chain 4:   Error evaluating the log probability at the initial value.
Chain 4: Exception: normal_lpdf: Scale parameter is -0.288136, but must be > 0!  (in 'model84027fb29ff_stan_test2' at line 38)

Chain 4: 
Chain 4: Gradient evaluation took 0 seconds
Chain 4: 1000 transitions using 10 leapfrog steps per transition would take 0 seconds.
Chain 4: Adjust your expectations accordingly!
Chain 4: 
Chain 4: 
Chain 4: Iteration:    1 / 2000 [  0%]  (Warmup)
Chain 4: Iteration:  200 / 2000 [ 10%]  (Warmup)
Chain 4: Iteration:  400 / 2000 [ 20%]  (Warmup)
Chain 4: Iteration:  600 / 2000 [ 30%]  (Warmup)
Chain 4: Iteration:  800 / 2000 [ 40%]  (Warmup)
Chain 4: Iteration: 1000 / 2000 [ 50%]  (Warmup)
Chain 4: Iteration: 1001 / 2000 [ 50%]  (Sampling)
Chain 4: Iteration: 1200 / 2000 [ 60%]  (Sampling)
Chain 4: Iteration: 1400 / 2000 [ 70%]  (Sampling)
Chain 4: Iteration: 1600 / 2000 [ 80%]  (Sampling)
Chain 4: Iteration: 1800 / 2000 [ 90%]  (Sampling)
Chain 4: Iteration: 2000 / 2000 [100%]  (Sampling)
Chain 4: 
Chain 4:  Elapsed Time: 1.233 seconds (Warm-up)
Chain 4:                1.091 seconds (Sampling)
Chain 4:                2.324 seconds (Total)
Chain 4: 
```

```{r}
print(m_hier,pars=c("gamma","tau","sigma"))
```

```{r}
mcmc_hier<-extract(m_hier)
str(mcmc_hier)
```

```{r}
#plot average response to explanatory variables
X_new<-model.matrix(~x+y,data=data.frame(x=seq(-2,2,by=0.2),y=0))
#get predicted values for each MCMC sample
pred_x1<-apply(mcmc_hier$gamma,1,function(beta) X_new %*% beta)
#now get median and 95% credible intervals
pred_x1<-apply(pred_x1,1,quantile,probs=c(0.025,0.5,0.975))
#same stuff for the second explanatory variables
X_new<-model.matrix(~x+y,data=data.frame(x=0,y=seq(-2,2,by=0.2)))
pred_x2<-apply(mcmc_hier$gamma,1,function(beta) X_new %*% beta)
pred_x2<-apply(pred_x2,1,quantile,probs=c(0.025,0.5,0.975))
```


```{r}
cols<-brewer.pal(10,"Set3")
par(mfrow=c(1,2),mar=c(4,4,0,1),oma=c(0,0,3,5))
plot(y~X[,2],pch=16,xlab="Temperature",ylab="Response variable",col=cols[id])
lines(seq(-2,2,by=0.2),pred_x1[1,],lty=2,col="red")
lines(seq(-2,2,by=0.2),pred_x1[2,],lty=1,lwd=3,col="blue")
lines(seq(-2,2,by=0.2),pred_x1[3,],lty=2,col="red")
plot(y~X[,3],pch=16,xlab="Nitrogen concentration",ylab="Response variable",col=cols[id])
lines(seq(-2,2,by=0.2),pred_x2[1,],lty=2,col="red")
lines(seq(-2,2,by=0.2),pred_x2[2,],lty=1,lwd=3,col="blue")
lines(seq(-2,2,by=0.2),pred_x2[3,],lty=2,col="red")
mtext(text = "Population-level response to the two\nexplanatory variables with 95% CrI",side = 3,line = 0,outer=TRUE)
legend(x=2.1,y=10,legend=paste("Gr",1:10),ncol = 1,col=cols,pch=16,bty="n",xpd=NA,title = "Group\nID")
```






```{r}
#now we could look at the variation in the regression coefficients between the groups doing caterpillar plots
ind_coeff<-apply(mcmc_hier$beta,c(2,3),quantile,probs=c(0.025,0.5,0.975))
df_ind_coeff<-data.frame(Coeff=rep(c("(Int)","X1","X2"),each=10),LI=c(ind_coeff[1,,1],ind_coeff[1,,2],ind_coeff[1,,3]),Median=c(ind_coeff[2,,1],ind_coeff[2,,2],ind_coeff[2,,3]),HI=c(ind_coeff[3,,1],ind_coeff[3,,2],ind_coeff[3,,3]))
gr<-paste("Gr",1:10)
df_ind_coeff$Group<-factor(gr,levels=gr)
#we may also add the population-level median estimate
pop_lvl<-data.frame(Coeff=c("(Int)","X1","X2"),Median=apply(mcmc_hier$gamma,2,quantile,probs=0.5))
ggplot(df_ind_coeff,aes(x=Group,y=Median))+geom_point()+
  geom_linerange(aes(ymin=LI,ymax=HI))+coord_flip()+
  facet_grid(.~Coeff)+
  geom_hline(data=pop_lvl,aes(yintercept=Median),color="blue",linetype="dashed")+
  labs(y="Regression parameters")
```

