---
title: Expected runs for every game state
author: Collin Erickson
date: '2019-05-11'
slug: expected-runs-for-every-game-state
categories: []
tags: [R, mlbgameday, baseball]
---

A state in a baseball game is defined by the number outs and which bases have baserunners.
We could include the count, inning, etc, but we're starting with the simplest version.
For a given state, there is an expected number of runs a team would score in
the remainder of that inning.
These tables are readily available online.
I want to try to recreate the tables using data from mlbgameday package.
It will require joining some tables and may be challenging.

```{r}
library(mlbgameday)
library(magrittr)
library(ggplot2)
library(dplyr)
```

Get data.

```{r, eval=F}
# Takes hours to get all the data for the year
dat <- get_payload(start = "2018-01-01", end = "2018-12-31")
```


```{r, include=F}
dat <- readRDS("C:/Users/cbe117/Documents/GitHub/website-hugo//scratch/MLB2018.rds")
dat$pitch$az <- as.numeric(dat$pitch$az)
dat$pitch$break_y <- as.numeric(dat$pitch$break_y)
dat$pitch$break_angle <- as.numeric(dat$pitch$break_angle)
```

The data for each at bat doesn't include where the bases runners start and end,
the pitch data has where the runners start

```{r}
dat$pitch %>% head
```

We're going to have to use runner to get the progress of base runners.

```{r}
dat$runner %>% head
```

First let's join atbat and pitch, same as before.

```{r}
# play_guid makes it only final pitch of each atbat
d2 <- inner_join(dat$pitch, dat$atbat, by=c("num", "gameday_link", "play_guid")) %>% 
  select(-des_es, atbat_des_es, event2_es, url.x, url.y)
```

Now we are going to join this with separate data frames for runners
that start on each base.
We'll have to make changes to the runner df before doing so.

```{r}
df1b <- dat$runner %>% filter(start=="1B") %>% mutate(on_1b=id, score_on_1b=score, on_1b_end=end) %>% 
  select(on_1b, score_on_1b, on_1b_end, num, gameday_link)
```

```{r}
inner_join(d2, df1b, c("num", "gameday_link", "on_1b")) %>% head
```

Looks okay, let's do all bases.

```{r}
df2b <- dat$runner %>% filter(start=="2B") %>% mutate(on_2b=id, score_on_2b=score, on_2b_end=end) %>% 
  select(on_2b, score_on_2b, on_2b_end, num, gameday_link)
df3b <- dat$runner %>% filter(start=="3B") %>% mutate(on_3b=id, score_on_3b=score, on_3b_end=end) %>% 
  select(on_3b, score_on_3b, on_3b_end, num, gameday_link)
```

```{r}
d2r1 <- left_join(d2, df1b, c("num", "gameday_link", "on_1b"))
d2r12 <- left_join(d2r1, df2b, c("num", "gameday_link", "on_2b"))
d2r123 <- left_join(d2r12, df3b, c("num", "gameday_link", "on_3b"))
```

I'm getting something weird where rows are duplicated and 
I don't think they should be.
I'll remove these.

```{r}
d3 <- d2r123[!duplicated(d2r123),]
d3 %>% head
```

To make sure d3 makes sense, check number of batters in each inning.

```{r}
d3 %>% group_by(inning.x, inning_side.x, gameday_link) %>% summarize(N=n()) %>% with(table(N))
```

The one with 38 batters in the inning is an error.
But for the most part these numbers look right,
so I'm not going to waste time on it.


Now we need to find the number of runs scored from before each play
until the end of the inning.
I'm not sure this is getting everything right, like if
a run is scored on the last play of the inning.

```{r}
runsatendofinning <- d3 %>% group_by(inning.x, inning_side.x, gameday_link) %>% summarize(runsatend=as.numeric(if(inning_side.x[1]=="top") {max(away_team_runs)} else {max(home_team_runs)}))
runsatendofinning %>% head
```
```{r}
runsatendofinning$runsatend %>% table
```

I don't think 25 runs were actually scored in an inning, this is likely an error I should fix.

Now we join this with the previous data.

```{r}
d4 <- inner_join(d3, runsatendofinning, c("inning.x", "inning_side.x", "gameday_link"))
d4 <- d4 %>% 
  mutate(runsfromthispointtoendofinning=runsatend - (ifelse(inning_side.x=="top", as.numeric(away_team_runs), as.numeric(home_team_runs))))
d4 %>% head
```

```{r}
d4$runsfromthispointtoendofinning %>% table
```

There's clearly some issue here since I'm getting that negative runs can be scored.


```{r}
d4 %>% filter(runsfromthispointtoendofinning>=0) %>% 
  mutate(someone_on_1b=!is.na(on_1b), someone_on_2b=!is.na(on_2b), someone_on_3b=!is.na(on_3b)) %>% 
  group_by(someone_on_1b, someone_on_2b, someone_on_3b, o) %>% summarize(meanruns=mean(runsfromthispointtoendofinning))
```

These numbers are wrong.
You can check the true values [here on FanGraphs](https://library.fangraphs.com/misc/re24/).
I think the problem is with the data.
The runs and outs seem to be from the end of each play, when I want the beginning.
And the negative runs is a big concern that I didn't fix.

## Why I'm failing

The runs and outs are shown after each event, but the base runners shown
are from before the event.
We want the state before each at bat, or after each at bat,
but they mixed between the two.

```{r}
d4$runbattingteam <- ifelse(d4$inning_side.x=="top", d4$away_team_runs, d4$home_team_runs)
d4 %>% filter(event=="Home Run") %>% with(table(as.numeric(runbattingteam)))
```

```{r}
d4 %>% filter(event=="Groundout") %>% with(table(as.numeric(o)))
```

Below we see that there are baserunners on home run plays,
meaning the base runners are from before the play.
```{r}
d4 %>% filter(event=="Home Run") %>% with(table(!is.na(on_1b)))
```

## But I can use on_xb_end?

```{r}
d4 %>% filter(runsfromthispointtoendofinning>=0) %>% 
  mutate(someone_on_1b_end=!is.na(on_1b_end), someone_on_2b_end=!is.na(on_2b_end), someone_on_3b_end=!is.na(on_3b_end)) %>% 
  group_by(someone_on_1b_end, someone_on_2b_end, someone_on_3b_end, o) %>% summarize(meanruns=mean(runsfromthispointtoendofinning))
```

No, this is terrible wrong.
`on_1b_end` is not saying that someone is on first at the end of the play,
it is saying where the player that started on first ended up.

```{r}
d4$on_1b_end %>% table
```

It doesn't even say "1B" if they don't leave the base? That's not helpful.

```{r}
d4$score_on_1b %>% table
```

Can I just calculate where they all end up?

```{r}
d4 %>% filter(runsfromthispointtoendofinning>=0) %>% 
    mutate(someone_on_1b_end=event=="Single" | (is.na(on_1b_end) & is.na(score_on_1b)),
           someone_on_2b_end=event=="Double" | (is.na(on_2b_end) & is.na(score_on_2b)) | (!is.na(on_1b_end) & on_1b_end=="2B"),
           someone_on_3b_end=event=="Triple" | (is.na(on_3b_end) & is.na(score_on_3b)) | (!is.na(on_2b_end) & on_2b_end=="3B") | (!is.na(on_1b_end) & on_1b_end=="3B")
    ) %>% 
    group_by(someone_on_1b_end, someone_on_2b_end, someone_on_3b_end, o) %>% summarize(meanruns=mean(runsfromthispointtoendofinning), N=n()) 
```

No this is a mess and terribly wrong.

## Conclusion

I failed to recreate expected runs based on the state (number of outs and positions of baserunners).
I got numbers that were off by a significant amount, but seemingly correlated with the true values.
The data has the baserunners from before the play, but the outs and runs from
after the play, making the data very hard to use.
It also doesn't say where the batter ended up on base.
Even with this data, I should have been able to figure it out, but it's not worth the effort now.
