---
title: Getting better MLB data from BaseballSavant in R
author: Collin Erickson
date: '2019-05-13'
slug: getting-better-mlb-data-from-baseballsavant-in-r
categories: []
tags: [R, baseball]
---

The MLB data I got using the R package mlbgameday
is kind of disappointing.
It doesn't have things like exit velocity,
and I saw last time that it's very hard to calculate
expected runs.
Google recommended the following article:
https://billpetti.github.io/2018-02-19-build-statcast-database-rstats/

I looked at it and it seems like it would allow me to get better data
than mlbgameday.
It has launch_angle, launch_speed, estimated wOBA, etc.
So here I'm going to try to follow the instructions in the post
to get a better data set.

First we load the packages.

```{r, eval=F}
require(readr)
require(dplyr)
require(xml2)
require(magrittr)
```

This will tell us which dates to check for.

```{r, eval=F}
dates_reduced <- read_csv("https://raw.githubusercontent.com/BillPetti/baseball_research_notebook/master/dates_statcast_build.csv")
```

We split those dates by year.

```{r, eval=F}
x2008season <- dates_reduced %>%
  filter(substr(GameDate, 1, 4) == 2008)

x2009season <- dates_reduced %>%
  filter(substr(GameDate, 1, 4) == 2009)

x2010season <- dates_reduced %>%
  filter(substr(GameDate, 1, 4) == 2010)

x2011season <- dates_reduced %>%
  filter(substr(GameDate, 1, 4) == 2011)

x2012season <- dates_reduced %>%
  filter(substr(GameDate, 1, 4) == 2012)

x2013season <- dates_reduced %>%
  filter(substr(GameDate, 1, 4) == 2013)

x2014season <- dates_reduced %>%
  filter(substr(GameDate, 1, 4) == 2014)

x2015season <- dates_reduced %>%
  filter(substr(GameDate, 1, 4) == 2015)

x2016season <- dates_reduced %>%
  filter(substr(GameDate, 1, 4) == 2016)

x2017season <- dates_reduced %>%
  filter(substr(GameDate, 1, 4) == 2017)
```

Here's the scraping function.

```{r, eval=F}
scrape_statcast_savant_pitcher_date <- function(start_date, end_date) {
  
  # extract year
  year <- substr(start_date, 1,4)
  
  # Base URL.
  url <- paste0("https://baseballsavant.mlb.com/statcast_search/csv?all=true&hfPT=&hfAB=&hfBBT=&hfPR=&hfZ=&stadium=&hfBBL=&hfNewZones=&hfGT=R%7C&hfC=&hfSea=",year,"%7C&hfSit=&player_type=pitcher&hfOuts=&opponent=&pitcher_throws=&batter_stands=&hfSA=&game_date_gt=",start_date,"&game_date_lt=",end_date,"&team=&position=&hfRO=&home_road=&hfFlag=&metric_1=&hfInn=&min_pitches=0&min_results=0&group_by=name&sort_col=pitches&player_event_sort=h_launch_speed&sort_order=desc&min_abs=0&type=details&")
  
  payload <- utils::read.csv(url)
  
  if (length(payload$pitch_type) > 0) {
  
  # Clean up formatting.
  payload$game_date <- as.Date(payload$game_date, "%Y-%m-%d")
  payload$des <- as.character(payload$des)
  payload$game_pk <- as.character(payload$game_pk) %>% as.numeric()
  payload$on_1b <- as.character(payload$on_1b) %>% as.numeric()
  payload$on_2b <- as.character(payload$on_2b) %>% as.numeric()
  payload$on_3b <- as.character(payload$on_3b) %>% as.numeric()
  payload$release_pos_x <- as.character(payload$release_pos_x) %>% as.numeric()
  payload$release_pos_z <- as.character(payload$release_pos_z) %>% as.numeric()
  payload$release_pos_y <- as.character(payload$release_pos_y) %>% as.numeric()
  payload$pfx_x <- as.character(payload$pfx_x) %>% as.numeric()
  payload$pfx_z <- as.character(payload$pfx_z) %>% as.numeric()
  payload$hc_x <- as.character(payload$hc_x) %>% as.numeric()
  payload$hc_y <- as.character(payload$hc_y) %>% as.numeric()
  payload$woba_denom <- as.character(payload$woba_denom) %>% as.numeric()
  payload$woba_value <- as.character(payload$woba_value) %>% as.numeric()
  payload$babip_value <- as.character(payload$babip_value) %>% as.numeric()
  payload$iso_value <- as.character(payload$iso_value) %>% as.numeric()
  payload$plate_z <- as.character(payload$plate_z) %>% as.numeric()
  payload$plate_x <- as.character(payload$plate_x) %>% as.numeric()
  payload$vx0 <- as.character(payload$vx0) %>% as.numeric()
  payload$vy0 <- as.character(payload$vy0) %>% as.numeric()
  payload$vz0 <- as.character(payload$vz0) %>% as.numeric()
  payload$ax <- as.character(payload$ax) %>% as.numeric()
  payload$ay <- as.character(payload$ay) %>% as.numeric()
  payload$az <- as.character(payload$az) %>% as.numeric()
  payload$sz_top <- as.character(payload$sz_top) %>% as.numeric()
  payload$sz_bot <- as.character(payload$sz_bot) %>% as.numeric()
  payload$hit_distance_sc <- as.character(payload$hit_distance_sc) %>% as.numeric()
  payload$launch_speed <- as.character(payload$launch_speed) %>% as.numeric()
  payload$launch_speed_angle <- as.character(payload$launch_speed_angle) %>% as.numeric()
  payload$launch_angle <- as.character(payload$launch_angle) %>% as.numeric()
  payload$estimated_ba_using_speedangle <- as.character(payload$estimated_ba_using_speedangle) %>% as.numeric()
  payload$estimated_woba_using_speedangle <- as.character(payload$estimated_woba_using_speedangle) %>% as.numeric()
  payload$effective_speed <- as.character(payload$effective_speed) %>% as.numeric()
  payload$release_speed <- as.character(payload$release_speed) %>% as.numeric()
  payload$zone <- as.character(payload$zone) %>% as.numeric()
  payload$release_spin_rate <- as.character(payload$release_spin_rate) %>% as.numeric()
  payload$release_extension <- as.character(payload$release_extension) %>% as.numeric()
  payload$barrel <- with(payload, ifelse(launch_angle <= 50 & launch_speed >= 98 & launch_speed * 1.5 - launch_angle >= 117 & launch_speed + launch_angle >= 124, 1, 0))
  payload$home_team <- as.character(payload$home_team)
  payload$away_team <- as.character(payload$away_team)
  
  return(payload)
  }
  
  else {
    vars <- names(payload)
    df <- lapply(vars, function(x) x <- NA)
    names(df) <- names(payload)
    payload_na <- bind_rows(df)
  
    return(payload_na)
    
    Sys.sleep(sample(x = runif(20, min = .01, max = 1), size = 1))
  }
}
```

First I'll test in a small data set.

```{r, eval=F}
x2008data <- x2008season[100:102,] %>%
  group_by(start_date) %>%
  do(scrape_statcast_savant_pitcher_date(.$start_date, .$end_date)) %>%
  ungroup() %>%
  select(-start_date)
```
```{r, eval=F}
x2008data
```


This looks okay, so I'll go and get the entire year.

```{r, eval=F}
x2008data <- x2008season %>%
  group_by(start_date) %>%
  do(scrape_statcast_savant_pitcher_date(.$start_date, .$end_date)) %>%
  ungroup() %>%
  select(-start_date)

(missing_2008 <- x2008data %>%
  filter(is.na(pitch_type)) %>%
  distinct(game_date) %>%
  select(game_date))
```

```{r, eval=F}
saveRDS(x2008data, "../../scratch/BaseballSavantData/MLB2008.rds")
```


```{r, eval=F}
x2009data <- x2009season %>%
  group_by(start_date) %>%
  do(scrape_statcast_savant_pitcher_date(.$start_date, .$end_date)) %>%
  ungroup() %>%
  select(-start_date)

gc()

(missing_2009 <- x2009data %>%
  filter(is.na(pitch_type)) %>%
  distinct(game_date) %>%
  select(game_date))


saveRDS(x2009data, "../../scratch/BaseballSavantData/MLB2009.rds")
rm(x2009data)
```



```{r, eval=F}
x2010data <- x2010season %>%
  group_by(start_date) %>%
  do(scrape_statcast_savant_pitcher_date(.$start_date, .$end_date)) %>%
  ungroup() %>%
  select(-start_date)

gc()

(missing_2010 <- x2010data %>%
  filter(is.na(pitch_type)) %>%
  distinct(game_date) %>%
  select(game_date))


saveRDS(x2010data, "./scratch/BaseballSavantData/MLB2010.rds")
rm(x2010data)
```


```{r, eval=F}
x2011data <- x2011season %>%
  group_by(start_date) %>%
  do(scrape_statcast_savant_pitcher_date(.$start_date, .$end_date)) %>%
  ungroup() %>%
  select(-start_date)

gc()

(missing_2011 <- x2011data %>%
  filter(is.na(pitch_type)) %>%
  distinct(game_date) %>%
  select(game_date))


saveRDS(x2011data, "./scratch/BaseballSavantData/MLB2011.rds")
rm(x2011data)
```

```{r, eval=F}
x2012data <- x2012season %>%
  group_by(start_date) %>%
  do(scrape_statcast_savant_pitcher_date(.$start_date, .$end_date)) %>%
  ungroup() %>%
  select(-start_date)

gc()

(missing_2012 <- x2012data %>%
  filter(is.na(pitch_type)) %>%
  distinct(game_date) %>%
  select(game_date))


saveRDS(x2012data, "../../scratch/BaseballSavantData/MLB2012.rds")
rm(x2012data)
```


```{r, eval=F}
x2013data <- x2013season %>%
  group_by(start_date) %>%
  do(scrape_statcast_savant_pitcher_date(.$start_date, .$end_date)) %>%
  ungroup() %>%
  select(-start_date)

gc()

(missing_2013 <- x2013data %>%
  filter(is.na(pitch_type)) %>%
  distinct(game_date) %>%
  select(game_date))


saveRDS(x2013data, "../../scratch/BaseballSavantData/MLB2013.rds")
rm(x2013data)
```


```{r, eval=F}
x2014data <- x2014season %>%
  group_by(start_date) %>%
  do(scrape_statcast_savant_pitcher_date(.$start_date, .$end_date)) %>%
  ungroup() %>%
  select(-start_date)

gc()

(missing_2014 <- x2014data %>%
  filter(is.na(pitch_type)) %>%
  distinct(game_date) %>%
  select(game_date))


saveRDS(x2014data, "./scratch/BaseballSavantData/MLB2014.rds")
rm(x2014data)
```


```{r, eval=F}
x2015data <- x2015season %>%
  group_by(start_date) %>%
  do(scrape_statcast_savant_pitcher_date(.$start_date, .$end_date)) %>%
  ungroup() %>%
  select(-start_date)

gc()

(missing_2015 <- x2015data %>%
  filter(is.na(pitch_type)) %>%
  distinct(game_date) %>%
  select(game_date))


saveRDS(x2015data, "../../scratch/BaseballSavantData/MLB2015.rds")
rm(x2015data)
```


```{r, eval=F}
x2016data <- x2016season %>%
  group_by(start_date) %>%
  do(scrape_statcast_savant_pitcher_date(.$start_date, .$end_date)) %>%
  ungroup() %>%
  select(-start_date)

gc()

(missing_2016 <- x2016data %>%
  filter(is.na(pitch_type)) %>%
  distinct(game_date) %>%
  select(game_date))


saveRDS(x2016data, "../../scratch/BaseballSavantData/MLB2016.rds")
rm(x2016data)
```


```{r, eval=F}
x2017data <- x2017season %>%
  group_by(start_date) %>%
  do(scrape_statcast_savant_pitcher_date(.$start_date, .$end_date)) %>%
  ungroup() %>%
  select(-start_date)

gc()

(missing_2017 <- x2017data %>%
  filter(is.na(pitch_type)) %>%
  distinct(game_date) %>%
  select(game_date))


saveRDS(x2017data, "../../scratch/BaseballSavantData/MLB2017.rds")
rm(x2017data)
```

I had issues with the years 2011, 2016, 2017.

## Fixing errors


### 2017 Error

The error for 2017:

```
Error: Column 'fielder_2' can't be converted from character to integer
In addition: There were 50 or more warnings (use warnings() to see the first 50)
```

To fix this I added the following into the function to convert all
the fielder values to integer.
After seeing that `fielder_2` is the exact same as `fielder_2.1` in all cases,
I just deleted the latter.

```{r, eval=F}
if ("fielder_2" %in% names(payload)) {
  
  payload$pitcher   <- as.character(payload$pitcher)   %>% as.integer
  payload$fielder_2 <- as.character(payload$fielder_2) %>% as.integer
  payload$fielder_3 <- as.character(payload$fielder_3) %>% as.integer
  payload$fielder_4 <- as.character(payload$fielder_4) %>% as.integer
  payload$fielder_5 <- as.character(payload$fielder_5) %>% as.integer
  payload$fielder_6 <- as.character(payload$fielder_6) %>% as.integer
  payload$fielder_7 <- as.character(payload$fielder_7) %>% as.integer
  payload$fielder_8 <- as.character(payload$fielder_8) %>% as.integer
  payload$fielder_9 <- as.character(payload$fielder_9) %>% as.integer
}
if (("fielder_2.1" %in% names(payload))) {
  payload$fielder_2.1 <- as.character(payload$fielder_2) %>% as.integer
}
```



### 2016 Error

The data from the 26th day, `"2016-04-28"` had an issue.
After rerunning it, I didn't get an error.
Maybe fixing 2017 fixed this too?

### 2011 Error

Also didn't have a problem when trying again.
Not sure what fixed it.


## 2018 data

The dates given above only included 2008 up through 2017.
To get the data, I did something like the following.
Turns out it's not really necessary to know the specific dates of games.

```{r, eval=F}
out <- list()
# Get the data, will include days with no data
for (i in 60:360) {
  datei <- as.Date("2018-01-01") + i
  print(c(i, datei))
  di <- scrape_statcast_savant_pitcher_date(datei, datei)
  out <- c(out, list(di))
}

# Remove days with no data, they have 1 row
outinds <- sapply(1:length(out), function(j) {nrow(out[[j]]) > 2})
out <- out[outinds]

out2 <- lapply(out, function(il) {il$fielder_2.1 <- as.integer(as.character(il$fielder_2.1)); il})
out3 <- bind_rows(out2)
if (any(out3$fielder2 != out3$fielder_2.1)) {stop("fielder2 not fielder2.1")}
out3$fielder_2.1 <- NULL
saveRDS(out3, "./scratch/BaseballSavantData/MLB2018.rds")
```


## Conclusion

Now I have better data, and for more years.
The new data has things like the score before and after each play,
launch angle and speed, and fielding alignment.
It looks like it has the runners on each base,
but not where they end up on each play, which
would be useful.
Overall it wasn't too hard to get the data following
the steps from Bill's blog,
there were only a few errors that could be resolved.
