---
title: On what counts do batters swing?
author: Collin Erickson
date: '2019-06-01'
slug: on-what-counts-do-batters-swing
categories: []
tags: [R, baseball]
---

Recently I've heard some things about when batters should swing.
Many hitters take the first pitch of an at bat to get a feel for the pitcher.
All hitters should take on 3-0.

I want to look at the rate at which batters swing based on the count.


```{r}
library(magrittr)
library(ggplot2)
library(dplyr)
```


Load the data as `d`

```{r, include=F}
d <- readRDS("C:/Users/cbe117/Documents/GitHub/website-hugo//scratch/BaseballSavantData/MLB2018.rds")
```

```{r, eval=F}
d <- readRDS("MLB2018.rds")
```

We can use `type` to see whether the batter put the ball into play.

```{r}
d$type %>% table
```

To see what qualifies as S vs X:

```{r}
d$description[d$type=="X"] %>% table
```

```{r}
d$description[d$type=="S"] %>% table
```

```{r}
d$description[d$type=="B"] %>% table
```

Now we will add a column for whether the batter swung or not.
They swung if it wasn't called a ball (B) or if it was an S
with description `called_strike`.

```{r}
d %<>% mutate(swung = !(type=="B" | description=="called_strike"))
table(d$swung, d$description)
```

I'll removed hit by pitch, but everything else looks okay.

```{r}
d %<>% filter(description!= "hit_by_pitch")
```

Now we will add a variable for the count.

```{r}
d %<>% mutate(count=paste(balls, strikes, sep='-'))
table(d$count)
```

I'll remove the cases with 4 balls, not sure where they come from.

```{r}
d %<>% filter(balls != 4) 
```

Now we can group by count and look at swing percentage.

```{r}
d %>% group_by(count) %>% summarize(swingpct = sum(swung) / n())
```

As expected, the swing rate is lowest on 3-0,
followed by 0-0.
It is highest on 3-2 by a significant margin,
but I don't have any good intuition on why
this should by higher than 2-2, 1-2, or 0-2.

