---
title: Using launch angle data
author: Collin Erickson
date: '2019-06-01'
slug: using-launch-angle-data
categories: []
tags: [R, baseball]
---

I'm going to look at some launch angle data
using the data scraped from Baseball Savant.

```{r}
library(magrittr)
library(ggplot2)
library(dplyr)
```


Load the data as `d`

```{r, include=F}
d <- readRDS("C:/Users/cbe117/Documents/GitHub/website-hugo//scratch/BaseballSavantData/MLB2018.rds")
```

```{r, eval=F}
d <- readRDS("MLB2018.rds")
```

```{r}
d %>% colnames()
```

It's not clear what `launch_speed_angle` is.

```{r}
d$launch_speed_angle %>% summary
```

`launch_speed` and `launch_angle` are as expected.
They both have a lot of missing values.
Most of these are pitches not hit.

```{r}
d$launch_speed %>% summary
```

```{r}
d$launch_angle %>% summary
```

Speed and angle are usually both false or true, which is good.

```{r}
table(is.na(d$launch_speed), is.na(d$launch_angle))
```

We'll filter to keep only pitches with launch angle and speed.

```{r}
dl <- d %>% filter(!is.na(launch_angle), !is.na(launch_speed))
```

We're going to look at at bat outcomes, shown in `events`.

```{r}
dl$events %>% table
```

`des` is useless and too specific.

```{r}
dl$des %>% head
```

A lot of the categories have few instances.
But we can see a clear pattern in `home_run` and extra base hits
compared to outs.

```{r}
ggplot(dl) + geom_point(aes(launch_speed, launch_angle)) + facet_wrap(. ~ events)
```

Estimated woba might be useful.
```{r}
dl$estimated_woba_using_speedangle %>% summary
```

```{r}
dl$estimated_woba_using_speedangle %>% is.na %>% summary
```

We will keep only events with at least 800 instances.
```{r}
ggplot(dl %>% group_by(events) %>% mutate(N=n()) %>% ungroup %>% filter(N>800)) + 
  geom_point(aes(launch_speed, launch_angle), alpha=.2) + facet_wrap(. ~ events)
```

Here's a plot for the densities of these.

```{r}
ggplot(dl %>% group_by(events) %>% mutate(N=n()) %>% ungroup %>% filter(N>800)) + 
  geom_density2d(aes(launch_speed, launch_angle)) + facet_wrap(. ~ events)
```

This plot below looks a little better.
There's a clear difference in what leads to extra-base hits compared to outs.
It's not clear what null is.
And there's nothing for a fly out.
```{r}
ggplot(dl %>% group_by(events) %>% mutate(N=n()) %>% ungroup %>% filter(N>800)) + 
  stat_density_2d(aes(launch_speed, launch_angle, fill=..level..), geom="polygon")+
  scale_fill_gradient(low="blue", high="red") + facet_wrap(. ~ events)
```

```{r}
dl %>% group_by(events) %>% mutate(N=n()) %>% ungroup %>% filter(N>800) %>% .$events %>% table
```

There's a lot of null events.
And they don't have descriptions either.
Could they be foul balls?
There were almost 130,000 foul balls in 2017, so 77,742 is a bit too low.
But it kind of makes sense from the plot above that there would be fewer
foul balls with a launch angle of zero since that would be going through
the bat.
The number of balls put in play is about right, but null doesn't
match with foul balls.
Maybe not all foul balls have recorded launch angle and speeds?

```{r}
(dl$events=="null") %>% table
```

I'm going to need to investigate this more,
but it seems like null might be foul balls.
