---
title: BABIP part 2
author: Collin Erickson
date: '2019-05-09'
slug: babip-part-2
categories: []
tags: [R, mlbgameday, baseball]
---

Last time I looked into BABIP to see if pitchers could control it.
I want to check a couple of more things.

I'm getting the same data as last time.

```{r}
library(mlbgameday)
library(magrittr)
library(ggplot2)
library(dplyr)
```


```{r, eval=F}
# Takes hours to get all the data for the year
dat <- get_payload(start = "2018-01-01", end = "2018-12-31")
```


```{r, include=F}
dat <- readRDS("C:/Users/cbe117/Documents/GitHub/website-hugo//scratch/MLB2018.rds")
dat$pitch$az <- as.numeric(dat$pitch$az)
dat$pitch$break_y <- as.numeric(dat$pitch$break_y)
dat$pitch$break_angle <- as.numeric(dat$pitch$break_angle)
```

Get pitcher names from at bat data.
`play_guid` is included since we
only want the final pitch of each at bat, not every pitch.
```{r}
d2 <- inner_join(dat$pitch, dat$atbat, by=c("num", "url", "play_guid"))
```

## Confidence intervals

So my idea is that I should check the confidence interval for each pitcher's
BABIP to see how much they overlap.

```{r}
inplayhits <- c("Double", "Single", "Triple")
inplayouts <- c("Bunt Groundout", "Bunt Lineout", "Bunt Pop Out", "Double Play", "Field Error", "Fielders Choice", "Fielders Choice Out", "Flyout", "Forceout", "Grounded Into DP", "Groundout", "Lineout", "Pop Out", "Triple Play")


BABIPbypitcher <- d2 %>% group_by(pitcher_name) %>% summarize(ipo=sum(event %in% inplayouts), iph=sum(event %in% inplayhits), N=iph+ipo, BABIP=iph/(iph+ipo)) %>% filter(N>100) %>% arrange(BABIP) %>% filter(!is.na(pitcher_name))
```

```{r}
BABIPbypitcher %<>% mutate(BABIP_low=BABIP - 2*sqrt(ipo*iph/N^3), BABIP_high=BABIP + 2*sqrt(ipo*iph/N^3)) 
```

Let's plot it.

```{r}
ggplot(data=BABIPbypitcher %>% mutate(i=1:n()), mapping=aes(BABIP, i)) + geom_line(aes(value, i, group=i), reshape2::melt(BABIPbypitcher %>% mutate(i=1:n()), id.vars="i", measure.vars=c("BABIP_low", "BABIP_high"))) + geom_point(color="yellow") + geom_vline(xintercept = sum(BABIPbypitcher$iph)/sum(BABIPbypitcher$N))
```

Okay, so each pitcher is a horizontal line and a yellow dot.
The yellow dot is their measured BABIP, the black line represents their 95% confidence interval.
We see that almost all of them cover the league average BABIP.
We can calculate what percentage of them include the league average.
We'd expect about 95% of them to include the league average value.

```{r}
leagueavgBABIP <- sum(BABIPbypitcher$iph)/sum(BABIPbypitcher$N)
BABIPbypitcher %<>% mutate(containsleagueavg = leagueavgBABIP > BABIP_low & leagueavgBABIP < BABIP_high)
BABIPbypitcher$containsleagueavg %>% table %>% print
(sum(BABIPbypitcher$containsleagueavg) / nrow(BABIPbypitcher)) %>% print
```

We get about 92.4%, which is pretty close to 95%.
It would be very unlikely this would be above 95%.
This is more convincing that pitchers have minimal control over BABIP than what I checked before.

### Test for proportions

What we are asking is whether a set of proportions are all equal
or if there is a difference.
Instead of making silly plots, I should try running a statistical test.
This is really easy with `prop.test`.

```{r}
prop.test(BABIPbypitcher$iph, BABIPbypitcher$N)
```

This prints way too much.
All we really want is the p-value.

```{r}
prop.test(BABIPbypitcher$iph, BABIPbypitcher$N)$p.value
```

The p value is less than .1%.
This would qualify as statistically significant.
But whenever we have this much data, almost anything that
does not exactly follow the null hypothesis
(equal proportions) will be significant.
It will be more useful to compare this to the values
calculated for other quantities.



I also want to see how the confidence intervals compare to other pitching numbers
to make sure it makes sense.
I'm going to find the number each pitcher had of some other common events
that are likely to vary more from pitcher to pitcher.


## Checking proportion of balls in play

```{r}
otherpitcherdata <- d2 %>% group_by(pitcher_name) %>% summarize(ipo=sum(event %in% inplayouts), iph=sum(event %in% inplayhits), ipall=ipo+iph, strikeout=sum(event %in% c("Strikeout", "Strikeout - DP")), walk=sum(event %in% c("Walk")), homerun=sum(event %in% c("Home Run")), allhits=iph+homerun, groundout=sum(event %in% c("Groundout", "Grounded Into DP", "Forceout", "Fielders Choice Out", "Fielders Choice")), flyout=sum(event %in% "Flyout"), N=n(), BABIP=iph/(iph+ipo)) %>% filter(N>100) %>% arrange(BABIP) %>% filter(!is.na(pitcher_name)) %>% mutate(i=1:n())
```



First let's check the proportion of at bats that get a ball in play.
This will be lower for strikeout/walk/home run pitchers.

```{r}
otherpitcherdata %>% mutate(ballinplayprop=(iph+ipo)/N, 
                            ballinplayprop_low=ballinplayprop-2*sqrt(ballinplayprop*(1-ballinplayprop)/N),
                            ballinplayprop_high=ballinplayprop+2*sqrt(ballinplayprop*(1-ballinplayprop)/N)
) %>% 
  {ggplot(data=., mapping=aes(ballinplayprop, i)) + geom_line(aes(value, i, group=i), reshape2::melt(., id.vars="i", measure.vars=c("ballinplayprop_low", "ballinplayprop_high"))) + geom_point(color="green") + geom_vline(xintercept = sum(.$iph+.$ipo)/sum(.$N))}
```

The proportion that includes the league mean is

```{r}
otherpitcherdata %>% mutate(ballinplayprop=(iph+ipo)/N, 
                            ballinplayprop_low=ballinplayprop-2*sqrt(ballinplayprop*(1-ballinplayprop)/N),
                            ballinplayprop_high=ballinplayprop+2*sqrt(ballinplayprop*(1-ballinplayprop)/N),
                            containsleagueavg = sum(.$iph+.$ipo)/sum(.$N) > ballinplayprop_low & sum(.$iph+.$ipo)/sum(.$N) < ballinplayprop_high
) %>% {table(.$containsleagueavg)}
```

This is just over half, or 52.9%.
Much less than 95%.
And we get a p-value of zero.

```{r}
otherpitcherdata %>% with(prop.test(iph+ipo, N)) %>% with(p.value)
```

This code is awful, and difficult to copy/paste to try other events.
I'll write a function for it.

## Function for comparing rates across pitchers

Here's a function that will do all the steps we need.

1. Get data for the events of interest `events1` as a proportion of `events1+events2`.
2. Make a plot with confidence intervals.
3. Find the confidence interval coverage.
4. Find p-value.

I'm requiring at least 20 of each events to avoid small sample issues.

```{r}
compareevents <- function(events1, events2=setdiff(unique(d2$event), events1)) {
  otherpitcherdata <- d2 %>% group_by(pitcher_name) %>% 
    summarize(e1=sum(event %in% events1), e2=sum(event %in% events2), 
              N=e1+e2, e1rate=e1/N, e1stderr=sqrt(e1*e2/N^3), 
              e1rate_low=e1rate-2*e1stderr, e1rate_high=e1rate+2*e1stderr) %>% 
    filter(e1>20, e2>20) %>% filter(!is.na(pitcher_name)) %>% 
    arrange(e1rate) %>% mutate(i=1:n()) 
  leagueavg <- sum(otherpitcherdata$e1)/sum(otherpitcherdata$N)
  otherpitcherdata %<>% mutate(containsleagueavg=e1rate_low < leagueavg & e1rate_high > leagueavg)
  
  # Make plot of CI
  # browser()
  print(ggplot(data=otherpitcherdata, 
               mapping=aes(e1rate, i)) + geom_line(aes(value, i, group=i),
                                                   reshape2::melt(otherpitcherdata, id.vars="i", measure.vars=c("e1rate_low", "e1rate_high"))) +
          geom_point(color="green") + 
          geom_vline(xintercept = leagueavg, color="red") +
          ggtitle(paste(c("Proportion of ", events1, "\nvs", events2),collapse = " "))
  )
  
  # Find coverage
  cat("Coverage is\n")
  print(table(otherpitcherdata$containsleagueavg))
  cat("\nProportion = ", sum(otherpitcherdata$containsleagueavg)/nrow(otherpitcherdata), '\n')
  
  # Find p-value
  cat("p-value is: ", with(otherpitcherdata, prop.test(e1, N)$p.value), "\n")
}
```

As a reminder, all of the events and their number of appearances is:
```{r}
table(d2$event)
```

Now we can quickly check how proportions vary across pitchers in the league for 2018.

### BABIP

We've already checked this, it's just a check to make sure the new function is correct.
It'll be a little different since now I require 20 hits and 20 outs instead
of just 100 in play events.

```{r}
compareevents(inplayhits, inplayouts)
```

### Strikeout proportion

Let's check strikeout proportion of all at bats.

```{r}
compareevents(c("Strikeout", "Strikeout - DP"))
```

Here there is clearly a difference across pitchers.
Again, this means that pitchers have control of this aspect of the game
and it's not random chance, like BABIP appears to be.

### Three true outcomes

```{r}
compareevents(c("Strikeout", "Strikeout - DP", "Walk", "Intent Walk", "Home Run"))
```

There is a difference across pitchers for the three true outcomes vs everything else (balls in play).


### Home runs vs other hits

```{r}
compareevents("Home Run", inplayhits)
```

Some pitchers give up a lot of homeruns.

### Ground ball outs vs fly ball outs

I'd rather check ground balls vs fly balls, including hits and outs,
but the various hits aren't split up by hit type.

```{r}
compareevents(c("Bunt Groundout", "Double Play", "Fielder Error", "Fielders Choice",
                "Fielders Choice Out", "Forceout", "Groundout"),
              c("Flyout", "Lineout", "Pop Out", "Sac Fly", "Sac Fly DP"))
```

This is significant, there is a difference between flyball and groundball pitchers.
About half of outs are fly balls, half are grounders.

### On base percentage

```{r}
compareevents(c("Home Run", inplayhits, "Walk", "Intent Walk", "Hit By Pitch"))
```

There is a difference across pitchers for their allowed on base percentage.


## Conclusion

I've looked again at some BABIP data.
I've added in confidence intervals to show that over 92% of the 95% confidence
intervals contain the league average, signififying that pitchers
don't have much control over BABIP.
p-values from proportion tests are useful for big samples,
but we looked at them to see how they differ.
We checked some other rates for pitchers, such as ground ball rate,
and everything else did show a difference across pitchers.

